---
title: "hmd_analysis"
author: '"PSTAT 296A"'
date: "`r Sys.Date()`"
output: pdf_document
---

```{r message=FALSE}
# import packages
library(gnm)
library(forecast)
library(StMoMo)
library(demography)
library(ggplot2)
library(tidyverse)
library(reshape2) # For data transformation
library(gridExtra)
```

```{r}
# load data
US_data <- hmd.mx(country="USA", username="amandahuang@ucsb.edu", password="Ilovepstat296!")
```

# Exploratory Analysis

## Age Profile Over Time

```{r}
# extract age and year vector
Ages <- US_data$age
Years <- US_data$year

# extract mortality rate matrices
Rates_Male <- US_data$rate$male
Rates_Female <- US_data$rate$female
Rates_Total <- US_data$rate$total

```

```{r fig.show='hold', out.width = "50%"}
# --- Plot 1: Total Mortality
plot(US_data, series="total", ylim = c(-10, 0), xlim = c(0,110), main = "USA: Total Mortality Rate")
# Add vertical dotted lines at age 20 and 60
abline(v = 25, col = "black", lty = 2, lwd = 1.5)   # age 25
abline(v = 65, col = "black", lty = 2, lwd = 1.5)   # age 65
legend("bottomright", legend=unique(Years),
  col=rainbow(length(Years)*1.25), ncol=8, pch=19,
  title="Year", cex=0.44)

# --- Plot 2: Male Mortality
plot(US_data, series="male",  ylim = c(-10, 0), main = "USA: Male Mortality Rate")
# Add vertical dotted lines at age 20 and 60
abline(v = 25, col = "black", lty = 2, lwd = 1.5)   # age 25
abline(v = 65, col = "black", lty = 2, lwd = 1.5)   # age 65
legend("bottomright", legend=unique(Years),
  col=rainbow(length(Years)*1.25), ncol=8, pch=19, 
  title="Year", cex=0.44)

# --- Plot 3: Female Mortality
plot(US_data, series = "female", ylim = c(-10, 0), main = "USA: Female Mortality Rate")
# Add vertical dotted lines at age 20 and 60
abline(v = 25, col = "black", lty = 2, lwd = 1.5)   # age 25
abline(v = 65, col = "black", lty = 2, lwd = 1.5)   # age 65
legend("bottomright", legend=unique(Years),
  col=rainbow(length(Years)*1.25), ncol=8, pch=19,
  title="Year", cex=0.44)

# --- Plot 4: Mean Mortality Profit
# Cal mean log-death rate across all year for each age
# strat sex
Alpha_Male <- mean(US_data, series = names(US_data$rate)[2], transform = TRUE, 
                   na.rm = TRUE)$y
Alpha_Female <- mean(US_data, series = names(US_data$rate)[1], transform = TRUE, 
                     na.rm = TRUE)$y
Alpha_Total <- mean(US_data, series = names(US_data$rate)[3], transform = TRUE, 
                    na.rm = TRUE)$y

alpha_df <- data.frame(Ages, Male = Alpha_Male, Female = Alpha_Female, 
                       Total = Alpha_Total)
# alpha_df

# modify dataset for plotting
alpha_long <- alpha_df %>%
  pivot_longer(
    cols = c(Male, Female, Total),       # columns to pivot
    names_to = "Gender",                 # new variable name
    values_to = "Mean_LogRate"                # value column name
  )
# alpha_long

# plot mean log-death rate
ggplot(alpha_long, aes(x = Ages, y = Mean_LogRate, color = Gender)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Figure 4. Mean Log-Mortality Rate by Age and Gender (a_x Profile)",
    x = "Age (x)",
    y = "Mean Log(Mortality Rate)"
  ) +
  theme_minimal()
```

These plots display the age specific log-mortality rate for every years from 1933 to 2023. Each line represents a single year, with the colors used to denote the year (e.g., the warm colors for early years, cool colors for recent years.)\
\textbf{Mortality Curve Shape:} All log-mortality curves show a J-shape: high mortality in infancy, started declining to a lowest point around age 10 in young childhood, then increased dramatically from age 15 to 20, followed by a continuous, near-exponential rise with age. In early years, some fluctuations at the older ages, reflecting data variability, probably due to lack of sample.\
\textbf{Mortality Improvement:} The mortality curve vertically shift downward over time, representing mortality improvement overtime. Also, the rate of change by age group is not uniform. Therefore, cohort effect was observed. \textbf{Gender Comparision:} The Female lines tend to be lower than the Male lines. From age 15 to 20, males mortality curve has a clear accident hump while it is less pronounced for female mortality. The accident hump is often due to lack of driving experience of young males and have higher crash rate. The effect is gender specific, tied to male mortality. Also, at the elder ages, male mortality have more variation than female mortality.\
\textbf{REMARK:} This study limit age interval from 25 to 65 because this age interval is where the opioid adjustment most likely take place combined with the analysis of missing data of opioid death. Also, limiting age group can the large variation at the from infancy age to young adulthood and elder ages, which can improve the modeling accuracy.

```{r}
# Define color 
col_all <- rainbow(length(Years)*1.25)  # default colors
recent_years <- 2016:2023                 # years to make black
col_all[Years %in% recent_years] <- "black"

# --- Total Mortality, 2016-2023
plot(US_data, series="total", ylim = c(-10, 0), xlim = c(0,110))
lines(US_data, series="total", ylim = c(-10, 0), year = 2016:2023, col = "black")
# Add vertical dotted lines at age 20 and 60
abline(v = 25, col = "black", lty = 2, lwd = 1.5)   # age 25
abline(v = 65, col = "black", lty = 2, lwd = 1.5)   # age 65
legend("bottomright", legend=unique(Years), col=col_all, ncol=8, pch=19,
  title="Year", cex=0.45)

```

From years 2016-2023, log death rate for age 20-45 shift up compared to previous years. This increase can be due to many event, including opioid death.

## Change in Mortality Over Time

```{r fig.show='hold', out.width = "50%"}
# change in mortality over years, age specific
plot(US_data, series = "total", plot.type = "time", xlab = "Year")

# Age group
plot(US_data, series = "total", plot.type = "time", xlab = "Year", age = 0:25,
     main = "USA: total death rates (1933-2023) Age: 0-24")
plot(US_data, series = "total", plot.type = "time", xlab = "Year", age = 26:40, 
     main = "USA: total death rates (1933-2023) Age: 26-40")
plot(US_data, series = "total", plot.type = "time", xlab = "Year", age = 41:75, 
     main = "USA: total death rates (1933-2023) Age: 41-75")
plot(US_data, series = "total", plot.type = "time", xlab = "Year", age = 76:100, 
     main = "USA: total death rates (1933-2023) Age: 76-100")
plot(US_data, series = "total", plot.type = "time", xlab = "Year", age = 101:110, 
     main = "USA: total death rates (1933-2023) Age: 101-110")

```

This graph displays the evolution of age-specific mortality over time, where each line represent an age group and shows how its log death rate has changed across year. And the warm color (red, orange) is younger ages and cool color (blue/ purple) represent older ages.

The overall pattern reveals a steady decrease in death rate across nearly all ages but with different gradient, indicating continuous mortality improvement over years. This downward movement reveal cohort effect. At year 2020 and after, death mortality increase across all ages but with different rate due to COVID.

From age 26-40, mortality rate has more fluctuation. Mortality rate show a notable decline from 1930s to 1970s, reaching a minimum around 1980. However, the mortality curve increase from 1980 to 1990 and more pronounced after 2010. This may correspond to behavioral or socioeconomic factors, like substance abuse.

From age 41-100, mortality rate shows a steady downward trend over time. From ages 101-110, the curves display considerable volatility probably due to limited data availability and sample size variation. The lack of a clear trend indicates that statistical uncertainty dominates at these extreme ages.

```{r}
# difference of log-mortality between gender
diff_data <- Rates_Male - Rates_Female

# plot delta(log-mort)
matplot(Years, t(diff_data), type = "l",
        col = adjustcolor(rainbow(nrow(diff_data)), alpha.f = 0.4),
        lwd = 0.8,
        ylab = "Male - Female (log mortality difference)",
        xlab = "Year",
        main = "Gender Difference in Log Mortality Rates (1933-2023)")
abline(h = 0, col = "black", lty = 2)  # reference line

```

This graph shows the difference between male and female log mortality rate across all ages over time, where each line represents a age. The curves above zero indicate higher male mortality while below zero indicate higher female mortality.

Most of the curves locate above zero, meaning male mortality rates have consistently been higher than female mortality rate across almost all age groups and time periods. But during the early age, such as infant time and child time, the mortality curves fluctuate around zero, indicating no gender advantage on mortality rate during that time.

After 1970s, the gender gap appears wider and reach its maximum from 1980 to 2000. Previously, we discover that total mortality for age 26-40 increase from 1980 to 2000. Combined with the increase of gender gap from 1980 to 2000, we suspect that for age group 26-40, from 1980 to 2000, the males mortality increase or decrease at a lower rate than females while females mortality rate decrease or at a higher rate.

```{r}
# Define gradient colors for each gender
male_colors   <- colorRampPalette(c("darkblue", "skyblue"))(length(26:40))
female_colors <- colorRampPalette(c("darkred", "pink"))(length(26:40))

plot(US_data, series = "male", plot.type = "time", age = 26:40, 
    col = male_colors, xlab = "Year", ylim = c(-8,-4.9), 
     main = "USA: male vs female death rates age: 26-40")
lines(US_data, series = "female", plot.type = "time", age = 26:40, 
      col = female_colors)
legend("topright",
       legend = c("Male (blue gradient)", "Female (red gradient)"),
       col = c("blue", "red"), lwd = 2, bty = "n")
```

The blue gradient represents males, while the red gradient represents females. Between 1980 and 1990, male death rate increase while females rate is stable and continue to decrease. Therefore, for age group 26-40, increase of mortality from 1980 to 1990 is gender specific to male.

In addition, for age group 26-40, the increase of mortality after 2010 cross whole population, regardless of gender, which could be potentially caused by drug overdose and opioid.

# Mortality Modeling

## Lee-Carter

```{r warning=FALSE, message=FALSE, fig.width=10, fig.height=10}
# Convert to StMoMoData
mort <- StMoMoData(US_data, type = "central", series = "total")

LC <- lc(link = "log", const = "sum")
LCfit <- fit(LC, data = mort, ages.fit = 25:65)

```

### CI for kappa
```{r}
# Construct (per-year) Gaussian CI for Lee–Carter kappa_t using RW-with-drift increments
# Returns a data frame: year, kappa, lower, upper
kappa_ci_df <- function(LCfit, alpha = 0.05, show_plot = TRUE) {
  # Extract kappa and years
  kappa <- as.numeric(LCfit$kt)
  years <- LCfit$years
  
  stopifnot(length(years) == length(kappa))
  if (length(kappa) < 2) stop("Need at least 2 years of kappa_t to estimate increments.")
  if (anyNA(kappa)) stop("kappa_t contains NA; please remove/impute before constructing CI.")
  
  # Increments (RW with drift): y_t = kappa_t - kappa_{t-1}
  y <- diff(kappa)
  
  # Drift estimate (not directly used in the per-year CI band below, but kept for completeness)
  theta_hat <- mean(y)
  
  # Innovations: omega_t = y_t - theta_hat
  omega_hat <- y - theta_hat
  
  # Innovation variance (MLE)
  sigma2_w_hat <- mean(omega_hat^2)
  sigma_w_hat  <- sqrt(sigma2_w_hat)
  
  # Gaussian CI (same width each year under this simple assumption)
  z <- qnorm(1 - alpha / 2)
  kappa_lower <- kappa - z * sigma_w_hat
  kappa_upper <- kappa + z * sigma_w_hat
  
  kt_df <- data.frame(
    year  = years,
    kappa = kappa,
    lower = kappa_lower,
    upper = kappa_upper
  )
  
  # plot
  p <- ggplot(kt_df, aes(x = year, y = kappa)) +
          geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
          geom_line() +
          labs(
            x = "Year",
            y = "kappa_t",
            title = paste0("GAPC Gaussian RW-based ",
               round((1 - alpha) * 100),
               "% CI for kappa_t (per year)")
          ) +
          theme_minimal()
  if (isTRUE(show_plot)) print(p) 
  
  return(kt_df)
}

kt_df <- kappa_ci_df(LCfit, alpha = 0.05)
```

### Parameter Plot

```{r}
# Plot LC ax, bx, kt with (per-year) kappa CI lines
# - Input: LCfit
# - Output: graphs (side effect)
# - Internally calls kappa_ci_df(LCfit, show_plot = FALSE) to get kt_df
plot_lc_components_with_kappa_ci <- function(LCfit) {
  # get CI df without plotting
  kt_df <- kappa_ci_df(LCfit, show_plot = FALSE)
  
  # robust year vector for kt plot (your original code uses colnames(LCfit$kt))
  kt_years <- as.numeric(colnames(LCfit$kt))
  if (anyNA(kt_years)) {
    # fallback if colnames are missing/non-numeric
    kt_years <- LCfit$years
  }
  
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar), add = TRUE)
  
  par(mfrow = c(3, 1), mar = c(5, 6, 3, 1))
  
  # ax
  plot(25:65, LCfit$ax, type = "l", lwd = 2, col = "black",
       main = expression(a[x] ~ "vs." ~ x),
       xlab = "Age", ylab = expression(a[x]),
       cex.main = 2, cex.lab = 2, cex.axis = 1.3)
  axis(1, at = seq(25, 65, by = 5), labels = seq(25, 65, by = 5), cex.axis = 1.3)
  
  # bx
  plot(25:65, LCfit$bx, type = "l", lwd = 2, col = "black",
       main = expression(beta[x]^{(1)} ~ "vs." ~ x),
       xlab = "Age", ylab = expression(beta[x]^{(1)}),
       cex.main = 2, cex.lab = 2, cex.axis = 1.3)
  axis(1, at = seq(25, 65, by = 5), labels = seq(25, 65, by = 5), cex.axis = 1.3)
  
  # kt with CI
  plot(kt_years, LCfit$kt[1, ], type = "l", lwd = 2, col = "black",
       main = expression(kappa[t]^{(1)} ~ "vs." ~ t),
       ylim = c(min(kt_df$lower), max(kt_df$upper)),
       xlab = "Year", ylab = expression(kappa[t]^{(1)}),
       cex.main = 2, cex.lab = 2, cex.axis = 1.3)
  
  lines(kt_df$year, kt_df$lower, col = "gray40", lty = 2, lwd = 2)
  lines(kt_df$year, kt_df$upper, col = "gray40", lty = 2, lwd = 2)
  
  legend("topright",
         legend = c("kappa(t)", "95% CI"),
         col = c("black", "gray40"),
         lty = c(1, 2),
         lwd = 2,
         bty = "n")
  
  invisible(NULL)
}
```

```{r warning=FALSE, message=FALSE, fig.width=10, fig.height=10}
plot_lc_components_with_kappa_ci(LCfit)
```

1)  Age Effect $a_x$

-   $a_x$ is monotone increasing smoothly and almost linearly across age 25 to 60, reflecting steadily rising baseline mortality with age on the log scale. And there is a noticeable jump at the last age around 65. Since the sample size around age 65 is relatively large and we need to group data above 65, this jump might be due to a boundary/grouping effect.

2)  Age Sensitivity to period index $\beta_x^{(1)}$

-   $b_x$ declines with age across 25-65, where higher values at younger adult ages and lower values at older adult ages. Also, a sudden drop at the last age around 65.
-   This means improvements or deterioration in the period index $\kappa_t$, produce larger proportional changes in mortality at younger ages and smaller proportional changes at older adult ages.\
-   This make sense because period-wide shocks like medical advances, epidemics, and economic conditions, have stronger impact on young and middle-age adult than at the older age.

3)  Period mortality index $\kappa_t$

-   $\kappa_t$ shows a strong downward trend over the long-run, implying substantial long-term mortality improvement for age 25-65. However, the decline is not constant or linear; instead, it occurs in distinct phases. The index falls sharply from approximately 1933–1955, representing the period of greatest mortality improvement. From 1955–1970, $\kappa_t$ remains relatively flat, indicating a temporary slowdown in improvements. After 1970, the decline resumes, reflecting a renewed period of improvement driven by medical advances. For recent years, around 2020 has a short-term mortality uptick, which is corresponding to COVID shock.
-   Because $\beta_x^{1}$ greater than $0$ for all ages here, a decrease in $\kappa_t$ corresponds to improvement in mortality across all ages.
-   A strong decline in 1930s to 1950s could be potentially due to post-war improvements in income and living standards, roll-out of public health measures, and improvements in maternal.
-   From 1980-2000, the United States experience a notable stagnation in adult mortality improvement. Heat disease mortality had been falling rapidly since the 1970s due to improved treatment and prevention but the rate of improvement slowed before accelerating again in 1990s which cardiac surgery improved and better hypertension and cholesterol control was introduced. Others factors include high prevalence of smoking among adults, rising obesity, HIV/AIDS epidemic which rose sharply from around 1983 to 1995.

### Goodness of Fit

```{r}
# Check goodness of fit: Lee-Carter
Dxt <- mort$Dxt     # death counts
Ext <- mort$Ext     # exposures
mxt_obs <- Dxt / Ext

# LC paramerter
ax <- as.numeric(LCfit$ax)        # alpha_x (vector over ages)
bx <- as.numeric(LCfit$bx[,1])    # beta_x
kt <- as.numeric(LCfit$kt[1,])    # kappa_t 

eta_xt <- outer(bx, kt, "*")
eta_xt <- sweep(eta_xt, 1, ax, "+")
# LC fitted mortality
mxt_fit <- exp(eta_xt)

# Extract ages to plot
ages_to_plot <- c(25, 30, 40, 50, 60, 65)
age_index <- match(ages_to_plot, mort$ages)
years <- mort$years

plot_list <- list()

for (i in seq_along(age_index)) {
  age <- ages_to_plot[i]
  idx <- age_index[i]
  
  df <- data.frame(
    year = years,
    observed = mxt_obs[idx, ],
    fitted   = mxt_fit[idx-25, ]
  )
  
  df_melt <- melt(df, id.vars = "year")
  
  p <- ggplot(df_melt, aes(x = year, y = value, color = variable)) +
    geom_line(size = 0.9) +
    scale_y_log10() + 
    labs(title = paste("Age", age, ": Observed vs Fitted LC Mortality"),
         y = "mortality (log scale)", x = "year",
         color = "") +
    theme_minimal()
  
  plot_list[[i]] <- p
}

grid.arrange(
  plot_list[[1]], plot_list[[2]], plot_list[[3]],
  plot_list[[4]], plot_list[[5]], plot_list[[6]],
  nrow = 2, ncol = 3
)
```

## RH(constant)
```{r warning=FALSE, message=FALSE, fig.width=10, fig.height=13, cache = TRUE}
# Renshaw and Haberman (Lee-Carter with cohorts)
# a_x + B_xk_t + gamma_t-x
# cohortAgeFun = "1", B_x^(0) = 1
mort_rh1 <- StMoMoData(US_data, type = "central", series = "total")
RH <- rh(link = "log", cohortAgeFun = "1")
RHfit <- fit(RH, data = mort_rh1, ages.fit = 25:65)
RHfit
resRH <- residuals(RHfit)        # deviance residuals

# parameters
plot(RHfit)                      # ax, bx, kt, gc
```

1)  Age Effect $a_x$ and Age Sensitivity to period index $\beta_x^{(1)}$

-   Baseline mortality $a_x$ and age sensitivity $\beta_x^{(1)}$ show a similar pattern as Lee-Carter model, where $a_x$ increases monotonically with age from 25 to 65 and $\beta_x^{(1)}$ declines with ages.

2)  Period mortality index $\kappa_t$

-   The period index also show a long-term downward trend consistent with declining mortality over time like Lee-Carter Model. However, the downward trend becomes smoother in RH compared to Lee-Carter because part of the temporal variation is now attributed to cohort effects.

3)  Cohort Effect $\gamma_{t-x}$

-   The cohort term shows a clear increasing trend by year of birth, where earlier cohorts born in the late 1800s to early 1900s have lower mortality, while more recent cohorts show higher mortality.

## RH (NP)

```{r warning=FALSE, message=FALSE, fig.width=10, fig.height=13}
# Renshaw and Haberman (Lee-Carter with cohorts)
# a_x + B_xk_t + B_0 gamma_t-x
# cohortAgeFun = "NP", B_x^(0) = NP
mort_rh2 <- StMoMoData(US_data, type = "central", series = "total")
RHnp_mod <- rh(link = "log", cohortAgeFun = "NP")
RHnp_fit <- fit(RHnp_mod, data = mort_rh2, ages.fit = 25:65)
```
### Parameter Plot
```{r}
# Plot RHnp fit components + kappa CI band (base R)
# - Input: RHnp_fit (a fitted StMoMo RHnp object)
# - Uses:  kappa_ci_df(LCfit, alpha=0.05, show_plot=FALSE) that you already defined
# - Output: produces the plots (returns invisibly a list with kt_df + layout info)

plot_RHnp_with_kappa_CI <- function(RHnp_fit, alpha = 0.05) {
  # --- get kappa CI df (no ggplot shown) ---
  kt_df <- kappa_ci_df(RHnp_fit, alpha = alpha, show_plot = FALSE)

  # --- layout (same as your code) ---
  layout_mat <- matrix(c(1, 0,
                         2, 3,
                         4, 5),
                       nrow = 3, byrow = TRUE)

  # keep old par settings so we can restore
  old_par <- par(no.readonly = TRUE)
  on.exit({
    layout(1)
    par(old_par)
  }, add = TRUE)

  layout(layout_mat)
  par(mar = c(5, 6, 3, 1))

  ages <- RHnp_fit$ages

  # 1) ax
  plot(ages, RHnp_fit$ax, type = "l", lwd = 2, col = "black",
       main = expression(a[x] ~ "vs." ~ x),
       xlab = "Age", ylab = expression(a[x]), cex.main = 2,
       cex.lab = 2, cex.axis = 1.3)
  axis(1, at = seq(min(ages), max(ages), by = 5), 
       labels = seq(min(ages), max(ages), by = 5), cex.axis = 1.3)

  # 2) bx (beta^(1)_x)
  plot(ages, RHnp_fit$bx, type = "l", lwd = 2, col = "black",
       main = expression(beta[x]^{(1)} ~ "vs." ~ x),
       xlab = "Age", ylab = expression(beta[x]^{(1)}), cex.main = 2,
       cex.lab = 2, cex.axis = 1.3)
  axis(1, at = seq(min(ages), max(ages), by = 5), 
       labels = seq(min(ages), max(ages), by = 5), cex.axis = 1.3)

  # 3) kt (kappa^(1)_t) with CI
  t_years <- as.numeric(colnames(RHnp_fit$kt))
  kappa1  <- as.numeric(RHnp_fit$kt[1, ])

  plot(t_years, kappa1, type = "l", lwd = 2, col = "black",
       main = expression(kappa[t]^{(1)} ~ "vs." ~ t),
       ylim = c(min(kt_df$lower), max(kt_df$upper)),
       xlab = "Year", ylab = expression(kappa[t]^{(1)}), cex.main = 2,
       cex.lab = 2, cex.axis = 1.3)

  lines(t_years, kt_df$lower, col = "gray40", lty = 2, lwd = 2)
  lines(t_years, kt_df$upper, col = "gray40", lty = 2, lwd = 2)

  legend("topright",
         legend = c("kappa(t)", "95% CI"),
         col = c("black", "gray40"),
         lty = c(1, 3),
         lwd = 2,
         bty = "n")

  # 4) b0x (beta^(2)_x in RH notation in your code)
  plot(ages, RHnp_fit$b0x, type = "l", lwd = 2, col = "black",
       main = expression(beta[x]^{(2)} ~ "vs." ~ x),
       xlab = "Age", ylab = expression(beta[x]^{(2)}), cex.main = 2,
       cex.lab = 2, cex.axis = 1.3)
  axis(1, at = seq(min(ages), max(ages), by = 5), 
       labels = seq(min(ages), max(ages), by = 5), cex.axis = 1.3)

  # 5) gc (gamma_{t-x} vs cohort)
  cohorts <- as.numeric(names(RHnp_fit$gc))
  plot(cohorts, RHnp_fit$gc, type = "l", lwd = 2, col = "black",
       main = expression(gamma[t-x] ~ "vs." ~ t-x),
       xlab = "Cohort", ylab = expression(gamma[t-x]), cex.main = 2,
       cex.lab = 2, cex.axis = 1.3)

  invisible(list(kt_df = kt_df))
}
```

```{r warning=FALSE, message=FALSE, fig.width=10, fig.height=13}
plot_RHnp_with_kappa_CI(RHnp_fit)
```

### Goodness of Fit

```{r}
# Ages and years actually used in the RH model
ages_fit  <- RHnp_fit$ages
years_fit <- RHnp_fit$years

# Extract deaths & exposures for those ages
Dxt_fit <- mort$Dxt[match(ages_fit, mort$ages), ]
Ext_fit <- mort$Ext[match(ages_fit, mort$ages), ]

# Observed mortality (central death rates)
mxt_obs_RH <- Dxt_fit / Ext_fit

# Fitted mortality from the RH model
mxt_fit_RH <- fitted(RHnp_fit, type = "rates")   # matrix age x year

ages_to_plot <- c(25, 30, 40, 50, 60, 65)       # all must be in ages_fit
age_index    <- match(ages_to_plot, ages_fit)

plot_list_RH <- vector("list", length(ages_to_plot))

for (i in seq_along(age_index)) {
  age <- ages_to_plot[i]
  idx <- age_index[i]

  df <- data.frame(
    year     = years_fit,
    observed = mxt_obs_RH[idx, ],
    fitted   = mxt_fit_RH[idx, ]
  )

  df_melt <- melt(df, id.vars = "year")

  p <- ggplot(df_melt, aes(x = year, y = value, color = variable)) +
    geom_line(size = 0.9) +
    scale_y_log10() +
    labs(title = paste("Age", age, ": Observed vs Fitted RH Mortality"),
         y = "mortality (log scale)", x = "year",
         color = "") +
    theme_minimal()

  plot_list_RH[[i]] <- p
}

grid.arrange(
  plot_list_RH[[1]], plot_list_RH[[2]], plot_list_RH[[3]],
  plot_list_RH[[4]], plot_list_RH[[5]], plot_list_RH[[6]],
  nrow = 2, ncol = 3
)
```

1)  Age Effect $a_x$

-   Baseline mortality $a_x$ remain a smooth monotonic increase across age 25-65 and a sharp increase around age 65, consistent with Lee-Carter model and the RH model with constant cohort loading.

2)  Age Sensitivity to period index $\beta_x^{(1)}$

-   In contrast to Lee–Carter and RH with constant cohort loading (both of which show a steady decline in $\beta_x^{(1)}$, the non-parametric RH model shows a different pattern: sensitivity decreases up to approximately age 60 and then increases again for the older ages in the range.
-   This suggests that mortality improvements over time are not uniformly strongest at younger ages, and that ages near 60 experience a renewed sensitivity to period mortality trends.

3)  Cohort Effect $\gamma_{t-x}$

-   The cohort index displays a substantially different pattern from the constant-loading RH model. Rather than a monotonic increase, the cohort effect declines for cohorts born between \~1880 and \~1940, and then increases sharply for later birth cohorts.
-   This reveals that the generational mortality disadvantage affecting recent cohorts is age-specific, not uniform across the adult age span.

4)  Cohort Age Loading $\beta_x^{(0)}(x)$

-   Unlike the constant-loading RH model where is $\beta_x^{(0)}$ flat, the non-parametric model shows clear variation across age. This indicates that cohort effects do not impact all ages equally—certain ages are more strongly influenced by generational effects than others.

## CBD Model

```{r, cache = TRUE}
mort_CBD <- StMoMoData(US_data, type = "initial", series = "total")
N <- 1e6
mort_CBD$Dxt <- round(mort_CBD$Dxt * N)
mort_CBD$Ext <- round(mort_CBD$Ext * N)

# Fit CBD
# Binomial Distribution
CBD <- cbd(link = "logit")
CBDfit <- fit(CBD, data = (mort_CBD), ages.fit = 25:65)
```

### CI for CBD
```{r}
# Construct (per-time) Gaussian CI for CBD kappa1_t and kappa2_t
# using RW-with-drift increments (same style as your kappa_ci_df).
#
# Inputs:
#   CBDfit    : fitted StMoMo cbd object (must contain $kt and $years)
#   alpha     : significance level (default 0.05 -> 95% CI)
#   show_plot : TRUE/FALSE; if TRUE, plot kappa1 and kappa2 with CI bands
#
# Outputs:
#   list(kt_df1 = ..., kt_df2 = ...)  (both data.frames with year,kappa,lower,upper)

kappa_ci_df_CBD <- function(CBDfit, alpha = 0.05, show_plot = TRUE) {

  # --- Extract kappa1, kappa2 and years ---
  if (is.null(CBDfit$kt)) stop("CBDfit$kt is missing.")
  if (is.null(CBDfit$years)) stop("CBDfit$years is missing.")

  kappa1 <- as.numeric(CBDfit$kt[1, ])
  kappa2 <- as.numeric(CBDfit$kt[2, ])
  years  <- CBDfit$years

  stopifnot(length(years) == length(kappa1), length(years) == length(kappa2))
  if (length(years) < 2) stop("Need at least 2 time points to estimate increments.")
  if (anyNA(kappa1) || anyNA(kappa2)) stop("kappa contains NA; please remove/impute before constructing CI.")

  # --- RW-with-drift increments ---
  y1 <- diff(kappa1)
  y2 <- diff(kappa2)

  theta_hat1 <- mean(y1)
  theta_hat2 <- mean(y2)

  omega_hat1 <- y1 - theta_hat1
  omega_hat2 <- y2 - theta_hat2

  # Innovation variance (MLE)
  sigma_w_hat1 <- sqrt(mean(omega_hat1^2))
  sigma_w_hat2 <- sqrt(mean(omega_hat2^2))

  # --- Gaussian CI (same-width band under this simple assumption) ---
  z <- qnorm(1 - alpha / 2)

  kappa_lower1 <- kappa1 - z * sigma_w_hat1
  kappa_upper1 <- kappa1 + z * sigma_w_hat1

  kappa_lower2 <- kappa2 - z * sigma_w_hat2
  kappa_upper2 <- kappa2 + z * sigma_w_hat2

  kt_df1 <- data.frame(
    year  = years,
    kappa = kappa1,
    lower = kappa_lower1,
    upper = kappa_upper1
  )

  kt_df2 <- data.frame(
    year  = years,
    kappa = kappa2,
    lower = kappa_lower2,
    upper = kappa_upper2
  )

  # --- Optional plots (base R, similar to what you did before) ---
  if (isTRUE(show_plot)) {
    old_par <- par(no.readonly = TRUE)
    on.exit(par(old_par), add = TRUE)

    par(mfrow = c(2, 1), mar = c(5, 6, 3, 1))

    # kappa1
    plot(kt_df1$year, kt_df1$kappa, type = "l", lwd = 2, col = "black",
         ylim = c(min(kt_df1$lower), max(kt_df1$upper)),
         main = expression(kappa[t]^{(1)} ~ "with 95% CI"),
         xlab = "Year", ylab = expression(kappa[t]^{(1)}),
         cex.main = 2, cex.lab = 2, cex.axis = 1.3)
    lines(kt_df1$year, kt_df1$lower, col = "gray40", lty = 2, lwd = 2)
    lines(kt_df1$year, kt_df1$upper, col = "gray40", lty = 2, lwd = 2)
    legend("topright",
           legend = c("kappa1(t)", "95% CI"),
           col = c("black", "gray40"),
           lty = c(1, 2),
           lwd = 2, bty = "n")

    # kappa2
    plot(kt_df2$year, kt_df2$kappa, type = "l", lwd = 2, col = "black",
         ylim = c(min(kt_df2$lower), max(kt_df2$upper)),
         main = expression(kappa[t]^{(2)} ~ "with 95% CI"),
         xlab = "Year", ylab = expression(kappa[t]^{(2)}),
         cex.main = 2, cex.lab = 2, cex.axis = 1.3)
    lines(kt_df2$year, kt_df2$lower, col = "gray40", lty = 2, lwd = 2)
    lines(kt_df2$year, kt_df2$upper, col = "gray40", lty = 2, lwd = 2)
    legend("topright",
           legend = c("kappa2(t)", "95% CI"),
           col = c("black", "gray40"),
           lty = c(1, 2),
           lwd = 2, bty = "n")

    par(mfrow = c(1, 1))
  }

  return(list(kt_df1 = kt_df1, kt_df2 = kt_df2))
}
```

### Parameter Plot with CI
```{r}
# Plot CBD kappa^(1) and kappa^(2) with CI (ggplot), with x-breaks built from data range
# - x_break_by controls the spacing of breaks (e.g., 5, 10, 20)
plot_CBD_kappa_CI_gg <- function(CBDfit, alpha = 0.05,
                                 x_break_by = 20,
                                 base_size = 12) {

  out <- kappa_ci_df_CBD(CBDfit, alpha = alpha, show_plot = FALSE)
  kt_df1 <- out$kt_df1
  kt_df2 <- out$kt_df2

  # Build breaks from data range
  year_min <- floor(min(c(kt_df1$year, kt_df2$year), na.rm = TRUE))
  year_max <- ceiling(max(c(kt_df1$year, kt_df2$year), na.rm = TRUE))
  x_breaks <- seq(year_min, year_max, by = x_break_by)

  p1 <- ggplot(kt_df1, aes(x = year, y = kappa)) +
    geom_line(color = "black", linewidth = 0.6) +
    geom_line(aes(y = lower), color = "gray40", linewidth = 0.6, linetype = "33") +
    geom_line(aes(y = upper), color = "gray40", linewidth = 0.6, linetype = "33") +
    labs(
      x = "year",
      y = expression(kappa[t]^{(1)}),
      title = expression(kappa[t]^{(1)} ~ "vs. t")
    ) +
    scale_x_continuous(breaks = x_breaks) +
    theme_classic(base_size = base_size) +
    theme(
      plot.title   = element_text(hjust = 0.5, size = 14),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 12),
      axis.text    = element_text(size = 10)
    )

  p2 <- ggplot(kt_df2, aes(x = year, y = kappa)) +
    geom_line(color = "black", linewidth = 0.6) +
    geom_line(aes(y = lower), color = "gray40", linewidth = 0.6, linetype = "33") +
    geom_line(aes(y = upper), color = "gray40", linewidth = 0.6, linetype = "33") +
    labs(
      x = "year",
      y = expression(kappa[t]^{(2)}),
      title = expression(kappa[t]^{(2)} ~ "vs. t")
    ) +
    scale_x_continuous(breaks = x_breaks) +
    theme_classic(base_size = base_size) +
    theme(
      plot.title   = element_text(hjust = 0.5, size = 14),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 12),
      axis.text    = element_text(size = 10)
    )

  gridExtra::grid.arrange(p1, p2, ncol = 1)

  invisible(list(p1 = p1, p2 = p2, kt_df1 = kt_df1, kt_df2 = kt_df2, x_breaks = x_breaks))
}
```

```{r}
plot_CBD_kappa_CI_gg(CBDfit, x_break_by = 20)
```

1)  $\kappa_t^{(1)}$: overall mortality leve in year $t$

-   $\kappa_t^{(1)}$ has a clear long-term decline from early 1930s to roughly 2010s, indicating substantial improvements in overall mortality over this period. The decline is not linear: the curve contains several stretches where it become almost flat, corresponding to periods in which mortality improvement temporarily slowed down. This pattern is consistent with the behavior of the period index $\kappa_t^{(1)}$ in both Lee-Carter and RH models.\
-   However, around 2010, CBD highlight that $\kappa_t^{(1)}$ stop declining and begins to increase, signalling a reversal in the long-run trend.

### Goodness of Fit

```{r}
# Ages and years actually used in the CBD model
ages_fit  <- CBDfit$ages
years_fit <- CBDfit$years

# Extract deaths & exposures for those ages
Dxt_fit <- mort_CBD$Dxt[match(ages_fit, mort_CBD$ages), ]
Ext_fit <- mort_CBD$Ext[match(ages_fit, mort_CBD$ages), ]

# Observed initial death probabilities q_{x,t}
qxt_obs_CBD <- Dxt_fit / Ext_fit

# Fitted q_{x,t} from CBD model
qxt_fit_CBD <- fitted(CBDfit, type = "rates")   # matrix: age x year

ages_to_plot <- c(25, 30, 40, 50, 60, 65)   # all must be in ages_fit
age_index    <- match(ages_to_plot, ages_fit)

plot_list_CBD <- vector("list", length(ages_to_plot))

for (i in seq_along(age_index)) {
  age <- ages_to_plot[i]
  idx <- age_index[i]

  df <- data.frame(
    year     = years_fit,
    observed = qxt_obs_CBD[idx, ],
    fitted   = qxt_fit_CBD[idx, ]
  )

  df_melt <- melt(df, id.vars = "year")

  p <- ggplot(df_melt, aes(x = year, y = value, color = variable)) +
    geom_line(size = 0.9) +
    scale_y_log10() +   # log scale for probabilities (optional but consistent)
    labs(title = paste("Age", age, ": Observed vs Fitted CBD Mortality"),
         y = "death probability (log scale)", x = "year",
         color = "") +
    theme_minimal()

  plot_list_CBD[[i]] <- p
}

grid.arrange(
  plot_list_CBD[[1]], plot_list_CBD[[2]], plot_list_CBD[[3]],
  plot_list_CBD[[4]], plot_list_CBD[[5]], plot_list_CBD[[6]],
  nrow = 2, ncol = 3
)
```

# Goodness of Fit

## Scaled Deviance Residuals

We use deviance residuals, which measure how much each observation contributes to the likelihood loss relative to a perfectly fitting model. 

$$
r_{xt} = \text{sign}(d_{xt}-\hat{d}_{xt})\sqrt{\frac{\text{dev}(xt)}{\hat{\phi}}
$$

$$
\hat{\phi} = \frac{D(d_{xt}, \hat{d_{xt}})}{K-v}
$$ 
- $D(d_{xt}, \hat{d}_{xt}) = \sum_x\sum_tw_{xt}dev(x,t)$ is total deviance of the model, 
- $K=\sum_x\sum_tw_{xt}$: number of observations (age-year cells) 
- $v$: effective number of estimated parameters 

### Deviance Contribution dev(x,t) GLM Deviance: 
$D = 2\{l(\text{saturated)}-l(\text{fitted})\}$ 
### Poisson Deviance Contribution Assume 
$$
D_{x,t} \sim \text{Poisson}(u_{x,t}), \quad u_{x,t} = \hat{D}_{x,t}
$$ 
Poisson log-likelihood: 
$$
l(u;y) = y\text{log}u-u
$$ 
- Saturated: $u=y$, $l(y;y)=ylogy-y$. 
- Fitted: $l(\hat{u};y) = y\text{log}\hat{u}-\hat{u}$ 
$$
dev(y,\hat{u}) = 2[(y\text{log}y-y)-(y\text{log}\hat{u}-\hat{u})]=2[y\text{log}(\frac{y}{\hat{u}})-(y-\hat{u})]
$$ 
For mortality cells: 
$$
dev(x,t) = 2[D_{x,t}\text{log}(\frac{D_{x,t}}{\hat{D}_{x,t}})-(D_{x,t}-\hat{D}_{x,t})]
$$ 
### Binomial Deviance Contribution 
Assume 
$$
D_{x,t} \sim Binomial(E_{x,t}^0, q_{x,t}), \quad \hat{D}_{x,t}=E_{x,t}^0\hat{q}_{x,t}
$$ 
Log-likelihood (again, constants cancel): 
$$
l(q;y) = y\text{log}q+(E-y)\text{log}(1-q)
$$ 
Saturated has $q = y/E$. Plugging into the deviance difference gives: 
$$
dev(y,\hat{q}) = 2[y\text{log}(\frac{y}{E\hat{q}}) + (E-y)\text{log}(\frac{E-y}{E(1-\hat{q})})
$$ 
For mortality cells, in terms of $\hat{D}=E\hat{q}$ 
$$
dev(x,t)) = 2[d_{xt}\text{log}(\frac{d_{xt}}{\hat{d}_{xt}}) + (E_{x,t}^0-d_{xt})\text{log}(\frac{E_{x,t}^0-d_{xt}}{E_{x,t}^0-\hat{d_{xt}}})]
$$ 
The total deviance of the model, $K=\sum_x\sum_tw_{xt}$ is the number of observations in the data and $v$ is the effective number of parameters in the model. 
$$
D(d_{xt}, \hat{d}_{xt}) = \sum_x\sum_tw_{xt}dev(x,t)
$$


```{r fig.show='hold', out.width = "50%"}
# scatter plots of residuals
# by age, period, and cohort
resid_LC <- residuals(LCfit)
plot(resid_LC, type = "scatter", main="LC Residuals")   

resid_RH2 <- residuals(RHnp_fit)
plot(resid_RH2, type = "scatter", main="RH Residuals") 

resid_CBD <- residuals(CBDfit)
plot(resid_CBD, type = "scatter", main="CBD Residuals") 
```

```{r fig.show='hold', out.width = "50%"}
# Residual heatmaps
plot(resid_LC,  type="colourmap", main="LC residuals")
plot(resid_RH2, type="colourmap", main="RH (NP) residuals")
plot(resid_CBD,type="colourmap", main="CBD residuals")
```

## AIC and BIC

Generally, models with more parameters provide a better fit to the data. To avoid over-parametrization and compare the relative performance of several models, it is common in the mortality literature to use information criteria which modify the maximum likelihood criterion by penalizing models with more parameters.

AIC and BIC compare models by trading off: - Goodness of fit - Number of parameters

AIC: 
$$
AIC = 2\nu - 2\mathcal{L}
$$ 
BIC: 
$$
BIC = \nu \log K - 2\mathcal{L}
$$ 
where: - $\mathcal{L}=$ maximized log-likelihood - $\nu=$ number of estimated parameters - $K$ = number of observation

```{r}
AIC(LCfit, RHnp_fit, CBDfit) %>%
  mutate(AIC = format(AIC, scientific = FALSE, big.mark = ","))

BIC(LCfit, RHnp_fit, CBDfit) %>%
  mutate(BIC = format(BIC, scientific = FALSE, big.mark = ","))
```

# Forecast

## Multivariate random walk with drfit

Define the $N$-vector 
$$
\kappa_t = \left( \kappa_t^{(1)}, \kappa_t^{(2)}, \dots, \kappa_t^{(N)} \right)^\top.
$$ 
Assume 
$$
\kappa_t = \delta + \kappa_{t-1} + \xi_t, \quad \xi_t \sim \mathcal{N}(0, \Sigma).
$$ 
This is a $\text{VARIMA}(0,1,0)$ with drift.

\textbf{h-step forecast (conditional mean)} Iterate: 
\textbf{Simulation form} 
$$
\kappa_{t+h} = \kappa_t + h\delta + \sum_{j=1}^h \xi_{t+j}.
$$ 
So simulate $\xi_{t+j} \sim \mathcal{N}(0, \Sigma)$ and accumulate.

Conditional on information up to time $t$ (call it $\mathcal{F}_t$), $\kappa_t$ is known, and the future noises are independent mean-zero Gaussians. 
$$
\mathbb{E}\left[ \kappa_{t+h} \mid \mathcal{F}_t \right] = \kappa_t + h\delta.
$$ 
\textbf{h-step forecast variance} Because increments add: 
$$
\text{Var}\left( \kappa_{t+h} \mid \mathcal{F}_t \right) = \text{Var}\left( \sum_{j=1}^h \Sigma\right) = h\Sigma.
$$

And since it's a linear combination of normals: 
$$
\kappa_{t+s} \mid \mathcal{F}_t \sim N(\kappa_t+s\delta, s\Sigma)
$$

\textbf{Estimating $\delta, \Sigma$ (simple, useful derivation):} 
Define increments $\Delta \kappa_t = \kappa_t - \kappa_{t-1}$. Under the model, 
$$
\Delta \kappa_t = \delta + \xi_t.
$$ 
So 
$$
\widehat{\delta} = \frac{1}{T-1} \sum_{t=2}^T \Delta \kappa_t, \quad \widehat{\Sigma} = \frac{1}{T-2} \sum_{t=2}^T \left( \Delta \kappa_t - \widehat{\delta} \right) \left( \Delta \kappa_t - \widehat{\delta} \right)^\top.
$$

## Univariate ARIMA for each $\kappa$

For each $i$,

$$
\Delta^d_i \kappa^{(i)}_t = \delta^{(i)} + \sum_{j=1}^{p_i} \phi^{(i)}_j \Delta^d_i \kappa^{(i)}_{t-j} + \sum_{j=1}^{q_i} \theta^{(i)}_j \varepsilon^{(i)}_{t-j} + \varepsilon^{(i)}_t, \quad \varepsilon^{(i)}_t \sim \mathcal{N}(0, \sigma^2_i).
$$

Then you use standard ARIMA forecasting formulas (via Kalman filter / innovations algorithm under the hood), producing:

\begin{itemize}
    \item forecast mean $\hat{\kappa}^{(i)}_{t+h}$
    \item forecast variance $\widehat{\text{Var}}(r^{(i)}_{t+h})$
\end{itemize}

Practical note: Approach 1 preserves correlation among period components via $\Sigma $.

## Rolling one-step-ahead algorithm

```{r}
get_fc_rates_vec <- function(fc, ages) {
  r <- fc$rates
  if (is.null(dim(r))) {
    v <- r
  } else {
    v <- r[, 1]
  }
  v <- as.numeric(v)
  names(v) <- rownames(as.matrix(fc$rates)) %||% names(fc$rates) # optional, but safer

  # best: just reorder by ages using names when available
  if (!is.null(names(fc$rates)) || !is.null(rownames(as.matrix(fc$rates)))) {
    nm <- if (!is.null(names(fc$rates))) names(fc$rates) else rownames(as.matrix(fc$rates))
    names(v) <- nm
    v <- v[as.character(ages)]
  }
  v
}

```

### LC
```{r}
## LC
# setting
evalYears <- 2003:2023
origins   <- evalYears - 1
ages      <- 25:65

mort_F <- StMoMoData(US_data, type = "central", series = "total")

# ── Extract observed data ONCE outside loop ───────────────────
obs_death <- mort_F$Dxt[as.character(ages), as.character(evalYears)]
obs_exp   <- mort_F$Ext[as.character(ages), as.character(evalYears)]
obs_mx    <- obs_death / obs_exp

# ── Forecast rate matrices ────────────────────────────────────
mx_hat_lc_rw <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                        dimnames = list(as.character(ages), as.character(evalYears)))
mx_hat_lc_ar <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                        dimnames = list(as.character(ages), as.character(evalYears)))

# ── CI matrices ───────────────────────────────────────────────
mx_hat_lc_rw_lower <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                              dimnames = list(as.character(ages), as.character(evalYears)))
mx_hat_lc_rw_upper <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                              dimnames = list(as.character(ages), as.character(evalYears)))
mx_hat_lc_ar_lower <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                              dimnames = list(as.character(ages), as.character(evalYears)))
mx_hat_lc_ar_upper <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                              dimnames = list(as.character(ages), as.character(evalYears)))

# ── LC CI extraction function ─────────────────────────────────
extract_lc_ci <- function(fitLC, fcLC, ages) {
  ax <- as.numeric(fitLC$ax[as.character(ages)])
  bx <- as.numeric(fitLC$bx[as.character(ages), 1])
  
  kt_mean  <- as.numeric(fcLC$kt.f$mean)
  kt_lower <- as.numeric(fcLC$kt.f$lower)
  kt_upper <- as.numeric(fcLC$kt.f$upper)
  
  log_mx_mean  <- ax + bx * kt_mean
  log_mx_upper <- ax + bx * ifelse(bx >= 0, kt_upper, kt_lower)
  log_mx_lower <- ax + bx * ifelse(bx >= 0, kt_lower, kt_upper)
  
  data.frame(
    age      = ages,
    mx_mean  = exp(log_mx_mean),
    mx_lower = exp(log_mx_lower),
    mx_upper = exp(log_mx_upper)
  )
}

# ── Forecast loop ─────────────────────────────────────────────
for (i in seq_along(origins)) {
  T <- origins[i]
  
  fitLC_f <- fit(LC, data = mort_F, ages.fit = 25:65, years.fit = 1933:T)
  print(paste("Year:", T, fitLC_f$conv))
  
  fcLC_rw <- forecast(fitLC_f, h = 1, level = 95)
  fcLC_ar <- forecast(fitLC_f, h = 1, kt.method = "iarima", 
                      kt.order = c(1, 0, 0), level = 95)
  
  # point forecast
  mx_hat_lc_rw[, as.character(T+1)] <- get_fc_rates_vec(fcLC_rw, ages)
  mx_hat_lc_ar[, as.character(T+1)] <- get_fc_rates_vec(fcLC_ar, ages)
  
  # CI
  ci_rw <- extract_lc_ci(fitLC_f, fcLC_rw, ages)
  ci_ar <- extract_lc_ci(fitLC_f, fcLC_ar, ages)
  
  mx_hat_lc_rw_lower[, as.character(T+1)] <- ci_rw$mx_lower
  mx_hat_lc_rw_upper[, as.character(T+1)] <- ci_rw$mx_upper
  mx_hat_lc_ar_lower[, as.character(T+1)] <- ci_ar$mx_lower
  mx_hat_lc_ar_upper[, as.character(T+1)] <- ci_ar$mx_upper
}

```

#### Save Output for LC
```{r}
# Save as CSV
write.csv(mx_hat_lc_rw,       file = "mx_hat_lc_rw.csv",       row.names = TRUE)
write.csv(mx_hat_lc_ar,       file = "mx_hat_lc_ar.csv",       row.names = TRUE)
write.csv(mx_hat_lc_rw_lower, file = "mx_hat_lc_rw_lower.csv", row.names = TRUE)
write.csv(mx_hat_lc_rw_upper, file = "mx_hat_lc_rw_upper.csv", row.names = TRUE)
write.csv(mx_hat_lc_ar_lower, file = "mx_hat_lc_ar_lower.csv", row.names = TRUE)
write.csv(mx_hat_lc_ar_upper, file = "mx_hat_lc_ar_upper.csv", row.names = TRUE)
write.csv(obs_death,          file = "obs_death.csv",           row.names = TRUE)
write.csv(obs_exp,            file = "obs_exp.csv",             row.names = TRUE)
write.csv(obs_mx,             file = "obs_mx.csv",              row.names = TRUE)

# Remove X prefix from column names when loading
load_matrix <- function(file) {
  m <- as.matrix(read.csv(file, row.names = 1, check.names = FALSE))
  return(m)
}

mx_hat_lc_rw       <- load_matrix("mx_hat_lc_rw.csv")
mx_hat_lc_ar       <- load_matrix("mx_hat_lc_ar.csv")
mx_hat_lc_rw_lower <- load_matrix("mx_hat_lc_rw_lower.csv")
mx_hat_lc_rw_upper <- load_matrix("mx_hat_lc_rw_upper.csv")
mx_hat_lc_ar_lower <- load_matrix("mx_hat_lc_ar_lower.csv")
mx_hat_lc_ar_upper <- load_matrix("mx_hat_lc_ar_upper.csv")
obs_death          <- load_matrix("obs_death.csv")
obs_exp            <- load_matrix("obs_exp.csv")
obs_mx             <- load_matrix("obs_mx.csv")
```

### RH
```{r}
# setting for forecast
ages     <- 25:65
years    <- 1933:2023
evalYears <- 2003:2023
origins  <- evalYears - 1

# list of matrix
sum_list <- vector("list", length(origins))
names(sum_list) <- as.character(origins)
# list of best matrix
sum_list_best <- vector("list", length(origins))
names(sum_list_best) <- as.character(origins)

# ── Initialize forecast + CI matrices ─────────────────────────
mx_hat_rh_ar       <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                              dimnames = list(as.character(ages), as.character(evalYears)))
mx_hat_rh_ar_lower <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                              dimnames = list(as.character(ages), as.character(evalYears)))
mx_hat_rh_ar_upper <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                              dimnames = list(as.character(ages), as.character(evalYears)))

# ── CI extraction function ────────────────────────────────────
extract_rh_ci <- function(fitRH, fcRH, ages_fit) {
  
  bx  <- as.numeric(fitRH$bx[as.character(ages_fit), 1])
  b0x <- as.numeric(fitRH$b0x[as.character(ages_fit)])
  
  # kt forecast bounds
  kt_mean  <- as.numeric(fcRH$kt.f$mean)
  kt_lower <- as.numeric(fcRH$kt.f$lower)
  kt_upper <- as.numeric(fcRH$kt.f$upper)
  
  # extract forecast year from colnames
  forecast_year <- as.numeric(colnames(fcRH$kt.f$mean))
  
  # cohort for each age
  new_cohort <- forecast_year - ages_fit
  
  # extract fitted gc for all ages
  gc_val <- sapply(new_cohort, function(cohort) {
    as.numeric(fitRH$gc[as.character(cohort)])
  })
  
  # for age 25 (new cohort) replace with forecast gc
  age25_idx         <- which(ages_fit == 25)
  gc_val[age25_idx] <- as.numeric(fcRH$gc.f$mean)
  
  # gc CI deviation — only age 25 has uncertainty
  gc_lower_dev <- numeric(length(ages_fit))
  gc_upper_dev <- numeric(length(ages_fit))
  gc_lower_dev[age25_idx] <- as.numeric(fcRH$gc.f$lower) - as.numeric(fcRH$gc.f$mean)
  gc_upper_dev[age25_idx] <- as.numeric(fcRH$gc.f$upper) - as.numeric(fcRH$gc.f$mean)
  
  # use fcRH$rates as mean directly
  mx_mean  <- as.numeric(fcRH$rates[as.character(ages_fit)])
  log_mean <- log(mx_mean)
  
  # CI from kt deviation + gc deviation (age 25 only)
  kt_dev_upper <- ifelse(bx >= 0, kt_upper - kt_mean, kt_lower - kt_mean)
  kt_dev_lower <- ifelse(bx >= 0, kt_lower - kt_mean, kt_upper - kt_mean)
  
  log_mx_upper <- log_mean + bx * kt_dev_upper + b0x * gc_upper_dev
  log_mx_lower <- log_mean + bx * kt_dev_lower + b0x * gc_lower_dev
  
  data.frame(
    age      = ages_fit,
    mx_mean  = mx_mean,
    mx_lower = exp(log_mx_lower),
    mx_upper = exp(log_mx_upper)
  )
}

# ── Forecast loop ─────────────────────────────────────────────
for (i in seq_along(origins)) {
  T <- origins[i]
  set.seed(296 + T)
  cat("\n==========================================\n")
  cat("Processing Origin Year:", T, "| Forecast Year:", T+1, "\n")
  cat("==========================================\n")
  
  max_try     <- 15
  ages_fit    <- 25:65
  years_fit   <- 1933:T
  cohorts_fit <- (min(years_fit) - max(ages_fit)):(max(years_fit) - min(ages_fit))
  
  mort_rh_for <- StMoMoData(US_data, type = "central", series = "total")
  RHnp_mod    <- rh(link = "log", cohortAgeFun = "NP")
  
  cn <- c(
    "conv", "fc_ok",
    paste0("ax_",    ages_fit),
    paste0("bx1_",   ages_fit),
    paste0("bx2_",   ages_fit),
    paste0("kt_",    years_fit),
    paste0("gc_",    cohorts_fit),
    paste0("mx_fc_", ages_fit)
  )
  
  sum_mat <- matrix(NA_real_, nrow = max_try, ncol = length(cn),
                    dimnames = list(paste0("iter_", 1:max_try), cn))
  sum_mat_best <- matrix(NA_real_, nrow = 1, ncol = length(cn),
                         dimnames = list("iter_1", cn))
  best_loglik <- -Inf
  best_fc     <- NULL
  
  for (j in seq_len(max_try)) {
    # 1) Try fitting
    fitRH_j <- tryCatch(
      fit(RHnp_mod, data = mort_rh_for,
          ages.fit  = ages_fit,
          years.fit = years_fit,
          control   = list(maxit = 200)),
      error = function(e) NULL
    )
    
    # 2) Fit error
    if (is.null(fitRH_j) || isTRUE(fitRH_j$fail)) {
      cat("\n--- Iteration:", j, "Model Fit Fail ---\n")
      sum_mat[j, "conv"]  <- 0
      sum_mat[j, "fc_ok"] <- 0
      next
    }
    
    # 3) Fill parameters
    sum_mat[j, paste0("ax_",  ages_fit)] <- as.numeric(fitRH_j$ax[as.character(ages_fit)])
    bx1 <- fitRH_j$bx
    bx2 <- fitRH_j$b0x
    sum_mat[j, paste0("bx1_", ages_fit)] <- as.numeric(bx1[as.character(ages_fit), 1])
    sum_mat[j, paste0("bx2_", ages_fit)] <- as.numeric(bx2[as.character(ages_fit)])
    ky <- as.integer(colnames(fitRH_j$kt))
    sum_mat[j, paste0("kt_",  ky)]       <- as.numeric(fitRH_j$kt[1, ])
    gc       <- fitRH_j$gc
    gc_names <- as.integer(names(gc))
    sum_mat[j, paste0("gc_",  gc_names)] <- as.numeric(gc)
    
    # 4) Check convergence
    conv_j <- isTRUE(fitRH_j$conv)
    sum_mat[j, "conv"] <- as.numeric(conv_j)
    if (!conv_j) {
      cat("\n--- Iteration:", j, "Does not converge ---\n")
      sum_mat[j, "fc_ok"] <- 0
      next
    }
    cat("\n--- Iteration:", j, "Converged ---\n")
    
    # 5) Try forecasting
    fcRH_j <- tryCatch(
      forecast(fitRH_j, h = 1, kt.method = "mrwd",
               gc.order = c(1, 1, 0), level = 95),
      error = function(e) NULL
    )
    if (is.null(fcRH_j)) {
      cat("\n--- Iteration:", j, "Forecast Fail ---\n")
      sum_mat[j, "fc_ok"] <- 0
      next
    }
    cat("\n--- Iteration:", j, "Forecast Success ---\n")
    
    # 6) Record forecast
    sum_mat[j, "fc_ok"] <- 1
    sum_mat[j, paste0("mx_fc_", ages_fit)] <-
      as.numeric(fcRH_j$rates[as.character(ages_fit)])
    
    # 7) Update best
    ll_j <- fitRH_j$loglik
    if (ll_j > best_loglik && !is.null(fcRH_j)) {
      best_loglik       <- ll_j
      best_fc           <- fcRH_j
      sum_mat_best[1, ] <- sum_mat[j, ]
      
      # point forecast
      mx_hat_rh_ar[, as.character(T+1)] <- get_fc_rates_vec(fcRH_j, ages_fit)
      
      # CI from best model only
      ci_rh <- extract_rh_ci(fitRH_j, fcRH_j, ages_fit)
      mx_hat_rh_ar_lower[, as.character(T+1)] <- ci_rh$mx_lower
      mx_hat_rh_ar_upper[, as.character(T+1)] <- ci_rh$mx_upper
      
      cat("*** New best loglik:", round(ll_j, 2), "***\n")
    }
  }
  
  sum_list[[as.character(T)]]      <- sum_mat
  sum_list_best[[as.character(T)]] <- sum_mat_best
}
```

The key addition is `extract_rh_ci()` which handles **both `kt` and `gc`** bounds — unlike LC which only has `kt`. Both `bx` and `b0x` signs are checked independently to correctly maximize the CI width.

In the code above, we create, `mx_hat_rh_ar`: the forecast rate 2003-2023 but 2017 and 2018 missing. `sum_list_best`: a list contain the best iteration for each year. `sum_list`: matrix of parameters and forecast rate for all iteration each year.

RH fitting can fail to converge in some rolling windows because cohort-based models create a higher-dimensional, weakly identified likelihood surface with potential local stationary points. Multiple restarts mitigate sensitivity to initial values by re-optimizing from different starting points and selecting the best converged likelihood solution.


#### Save RH output
```{r}
# ── Save RH forecast matrices ──────────────────────────────────
saveRDS(mx_hat_rh_ar,       file = "mx_hat_rh_ar.rds")
saveRDS(mx_hat_rh_ar_lower, file = "mx_hat_rh_ar_lower.rds")
saveRDS(mx_hat_rh_ar_upper, file = "mx_hat_rh_ar_upper.rds")

# ── Save RH summary lists ──────────────────────────────────────
saveRDS(sum_list,      file = "sum_list_rh.rds")
saveRDS(sum_list_best, file = "sum_list_best_rh.rds")

# ── Save as CSV for easy viewing ──────────────────────────────
write.csv(mx_hat_rh_ar,       file = "mx_hat_rh_ar.csv",       row.names = TRUE)
write.csv(mx_hat_rh_ar_lower, file = "mx_hat_rh_ar_lower.csv", row.names = TRUE)
write.csv(mx_hat_rh_ar_upper, file = "mx_hat_rh_ar_upper.csv", row.names = TRUE)

# ── Verify all saved ──────────────────────────────────────────
cat("Files saved:\n")
cat(" mx_hat_rh_ar:      ", file.exists("mx_hat_rh_ar.rds"),       "\n")
cat(" mx_hat_rh_ar_lower:", file.exists("mx_hat_rh_ar_lower.rds"), "\n")
cat(" mx_hat_rh_ar_upper:", file.exists("mx_hat_rh_ar_upper.rds"), "\n")
cat(" sum_list_rh:        ", file.exists("sum_list_rh.rds"),        "\n")
cat(" sum_list_best_rh:   ", file.exists("sum_list_best_rh.rds"),   "\n")
```

```{r}
# Load Saved RH output
# ── Load RH forecast matrices ──────────────────────────────────
mx_hat_rh_ar       <- readRDS("mx_hat_rh_ar.rds")
mx_hat_rh_ar_lower <- readRDS("mx_hat_rh_ar_lower.rds")
mx_hat_rh_ar_upper <- readRDS("mx_hat_rh_ar_upper.rds")

# ── Load RH summary lists ──────────────────────────────────────
sum_list      <- readRDS("sum_list_rh.rds")
sum_list_best <- readRDS("sum_list_best_rh.rds")
```

#### Plot Parameters

```{r}
pdf("RH_parameter_plots.pdf", width = 40, height = 4 * length(years))

ages <- 25:65
status_colors <- c(
  "Converged + Forecast"        = "darkgreen",
  "Converged + No Forecast"     = "orange",
  "Not Converged + No Forecast" = "red"
)

years <- names(sum_list)
n_year <- length(years)

# Layout: rows = years, columns = 6 parameter blocks
par(mfrow = c(n_year, 6), mar = c(2,2,2,1))


# ---------- LOOP THROUGH YEARS ----------
for (y in years) {

  M <- as.data.frame(sum_list[[y]])

  # Drop rows where parameter part all NA
  M <- M[rowSums(!is.na(M[, -c(1,2)])) > 0, ]

  # Convert conv / fc_ok to logical
  M$conv  <- as.integer(M$conv) == 1
  M$fc_ok <- as.integer(M$fc_ok) == 1

  # Status label
  M$status <- with(M,
    ifelse(conv & fc_ok, "Converged + Forecast",
    ifelse(!conv & !fc_ok, "Not Converged + No Forecast",
           "Converged + No Forecast"))
  )

  # transparent colors by status
  cols <- adjustcolor(status_colors[M$status], alpha.f = 0.35)
  
  M$iter <- seq_len(nrow(M))

  # Extract parameter blocks (robust)
  ax_block  <- M[, grep("^ax_",     colnames(M)), drop = FALSE]
  bx1_block <- M[, grep("^bx1_",    colnames(M)), drop = FALSE]
  bx2_block <- M[, grep("^bx2_",    colnames(M)), drop = FALSE]
  kt_block  <- M[, grep("^kt_",     colnames(M)), drop = FALSE]
  gc_block  <- M[, grep("^gc_",     colnames(M)), drop = FALSE]
  mx_block  <- M[, grep("^mx_fc_",  colnames(M)), drop = FALSE]

  cols <- status_colors[M$status]

  # =========================
  # 5) PLOT BLOCKS
  # =========================

  # ax
  if (ncol(ax_block) > 0) {
    matplot(ages, t(ax_block),
            type = "l", lty = 1, col = cols,
            main = paste("ax -", y),
            xlab = "", ylab = "")
    legend("bottomright", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else plot.new()

  # bx1
  if (ncol(bx1_block) > 0) {
    matplot(ages, t(bx1_block),
            type = "l", lty = 1, col = cols,
            main = paste("bx1 -", y),
            xlab = "", ylab = "")
    legend("topright", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else plot.new()

  # bx2
  if (ncol(bx2_block) > 0) {
    matplot(ages, t(bx2_block),
            type = "l", lty = 1, col = cols,
            main = paste("bx2 -", y),
            xlab = "", ylab = "")
    legend("topright", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else plot.new()

  # kt (period index)
  if (ncol(kt_block) > 0) {
    kt_years <- as.numeric(sub("kt_", "", colnames(kt_block)))
    matplot(kt_years, t(kt_block),
            type = "l", lty = 1, col = cols,
            main = paste("kt -", y),
            xlab = "", ylab = "")
    legend("topright", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else plot.new()

  # gc (cohort index)
  if (ncol(gc_block) > 0) {
    gc_vals <- as.numeric(sub("gc_", "", colnames(gc_block)))
    matplot(gc_vals, t(gc_block),
            type = "l", lty = 1, col = cols,
            main = paste("gc -", y),
            xlab = "", ylab = "")
    legend("bottomright", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else plot.new()

  # mx forecast (only successful)
  idx_ok <- which(M$status == "Converged + Forecast")

  if (length(idx_ok) > 0 && ncol(mx_block) > 0) {
    matplot(ages, t(mx_block[idx_ok, , drop = FALSE]),
            type = "l", lty = 1,
            col = status_colors["Converged + Forecast"],
            main = paste("mx_fc -", as.integer(y) + 1),
            xlab = "", ylab = "")
    legend("topleft", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else {
    plot.new()
  }
}

# Reset layout
par(mfrow = c(1,1))
dev.off()
```

#### Plot for best fit and forecast

```{r}
pdf("RH_parameter_plots_best.pdf", width = 26, height = 4 * length(years)) 

ages <- 25:65
status_colors <- c(
  "Converged + Forecast"        = "darkgreen",
  "Converged + No Forecast"     = "orange",
  "Not Converged + No Forecast" = "red"
)


years <- names(sum_list_best)
# keep only years where there is at least ONE non-NA beyond conv/fc_ok
years_ok <- years[sapply(years, function(y) {
  M <- as.data.frame(sum_list_best[[y]])
  nrow(M) > 0 && any(!is.na(M[1, -c(1,2)]))
})]

years <- years_ok
n_year <- length(years)

# Layout: rows = years, columns = 6 parameter blocks
par(mfrow = c(n_year, 6), mar = c(2,2,2,1))


# ---------- LOOP THROUGH YEARS ----------
for (y in years) {
  
  M <- as.data.frame(sum_list_best[[y]])
  
  # Drop rows where parameter part all NA
  M <- M[rowSums(!is.na(M[, -c(1,2)])) > 0, ]
  
  # Convert conv / fc_ok to logical
  M$conv  <- as.integer(M$conv) == 1
  M$fc_ok <- as.integer(M$fc_ok) == 1
  
  # Status label
  M$status <- with(M,
                   ifelse(conv & fc_ok, "Converged + Forecast",
                          ifelse(!conv & !fc_ok, "Not Converged + No Forecast",
                                 "Converged + No Forecast"))
  )
  
  # transparent colors by status
cols <- adjustcolor(status_colors[M$status], alpha.f = 0.35)

  M$iter <- seq_len(nrow(M))
  
  #Extract parameter blocks (robust)
  # -------------------------
  ax_block  <- M[, grep("^ax_",     colnames(M)), drop = FALSE]
  bx1_block <- M[, grep("^bx1_",    colnames(M)), drop = FALSE]
  bx2_block <- M[, grep("^bx2_",    colnames(M)), drop = FALSE]
  kt_block  <- M[, grep("^kt_",     colnames(M)), drop = FALSE]
  gc_block  <- M[, grep("^gc_",     colnames(M)), drop = FALSE]
  mx_block  <- M[, grep("^mx_fc_",  colnames(M)), drop = FALSE]
  
  cols <- status_colors[M$status]
  
  # =========================
  # 5) PLOT BLOCKS
  # =========================
  
  # ax
  if (ncol(ax_block) > 0) {
    matplot(ages, t(ax_block),
            type = "l", lty = 1, col = cols,
            main = paste("ax -", y),
            xlab = "", ylab = "")
    legend("bottomright", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else plot.new()
  
  # bx1
  if (ncol(bx1_block) > 0) {
    matplot(ages, t(bx1_block),
            type = "l", lty = 1, col = cols,
            main = paste("bx1 -", y),
            xlab = "", ylab = "")
    legend("topright", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else plot.new()
  
  # bx2
  if (ncol(bx2_block) > 0) {
    matplot(ages, t(bx2_block),
            type = "l", lty = 1, col = cols,
            main = paste("bx2 -", y),
            xlab = "", ylab = "")
    legend("topright", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else plot.new()
  
  # kt (period index)
  if (ncol(kt_block) > 0) {
    kt_years <- as.numeric(sub("kt_", "", colnames(kt_block)))
    matplot(kt_years, t(kt_block),
            type = "l", lty = 1, col = cols,
            main = paste("kt -", y),
            xlab = "", ylab = "")
    legend("topright", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else plot.new()
  
  # gc (cohort index)
  if (ncol(gc_block) > 0) {
    gc_vals <- as.numeric(sub("gc_", "", colnames(gc_block)))
    matplot(gc_vals, t(gc_block),
            type = "l", lty = 1, col = cols,
            main = paste("gc -", y),
            xlab = "", ylab = "")
    legend("bottomright", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else plot.new()
  
  # mx forecast (only successful)
  idx_ok <- which(M$status == "Converged + Forecast")
  
  if (length(idx_ok) > 0 && ncol(mx_block) > 0) {
    matplot(ages, t(mx_block[idx_ok, , drop = FALSE]),
            type = "l", lty = 1,
            col = status_colors["Converged + Forecast"],
            main = paste("mx_fc -", as.integer(y) + 1),
            xlab = "", ylab = "")
    legend("topleft", legend=names(status_colors),
           col=status_colors, lty=1, cex=1, bty="n")
  } else {
    plot.new()
  }
}

# Reset layout
par(mfrow = c(1,1))

dev.off()
```

### CBD
```{r}
# ── Settings FIRST ────────────────────────────────────────────
ages      <- 25:65
years     <- 1933:2023
evalYears <- 2003:2023
origins   <- evalYears - 1

# ── CBD data ──────────────────────────────────────────────────
mort_CBD_for <- StMoMoData(US_data, type = "initial", series = "total")
obs_exp_CBD  <- mort_CBD_for$Ext[as.character(ages), as.character(evalYears)]

N <- 1e6
mort_CBD_for$Dxt <- round(mort_CBD_for$Dxt * N)
mort_CBD_for$Ext <- round(mort_CBD_for$Ext * N)
CBD <- cbd(link = "logit")

# ── CBD CI extraction function ────────────────────────────────
extract_cbd_ci <- function(fitCBD, fcCBD, ages) {
  
  xbar <- mean(ages)  # CBD uses mean-centered age
  
  # ── Extract kt forecast: mean, lower, upper ──────────────────
  # CBD has two kappas: k1 and k2
  kt_mean  <- fcCBD$kt.f$mean    # [2 x 1] matrix
  kt_lower <- fcCBD$kt.f$lower   # [2 x 1 x 1] array
  kt_upper <- fcCBD$kt.f$upper   # [2 x 1 x 1] array
  
  k1_mean  <- as.numeric(kt_mean[1,  ])
  k2_mean  <- as.numeric(kt_mean[2,  ])
  k1_lower <- as.numeric(kt_lower[1, ,])
  k2_lower <- as.numeric(kt_lower[2, ,])
  k1_upper <- as.numeric(kt_upper[1, ,])
  k2_upper <- as.numeric(kt_upper[2, ,])
  
  # ── For each age, compute logit(q) bounds ────────────────────
  # logit(q_x) = k1 + (x - xbar) * k2
  # coefficient of k1 = 1 (always positive)
  # coefficient of k2 = (x - xbar), sign depends on age vs xbar
  
  logit_mean  <- numeric(length(ages))
  logit_lower <- numeric(length(ages))
  logit_upper <- numeric(length(ages))
  
  for (idx in seq_along(ages)) {
    x    <- ages[idx]
    coef <- x - xbar  # coefficient for k2
    
    # k1: coefficient is always +1, so upper k1 → upper logit
    # k2: coefficient sign depends on (x - xbar)
    logit_mean[idx]  <- k1_mean  + coef * k2_mean
    
    logit_upper[idx] <- k1_upper + coef * ifelse(coef >= 0, k2_upper, k2_lower)
    logit_lower[idx] <- k1_lower + coef * ifelse(coef >= 0, k2_lower, k2_upper)
  }
  
  
  # convert back to probability scale to match fcCBD$rates
  inv_logit <- function(x) exp(x) / (1 + exp(x))
  
  data.frame(
    age      = ages,
    qx_mean  = inv_logit(logit_mean),
    qx_lower = inv_logit(logit_lower),
    qx_upper = inv_logit(logit_upper)
  )
}

# ── Initialize matrices (AFTER ages and evalYears defined) ────
qx_hat_CBD <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                     dimnames = list(as.character(ages), as.character(evalYears)))

qx_hat_CBD_lower <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                            dimnames = list(as.character(ages), as.character(evalYears)))

qx_hat_CBD_upper <- matrix(NA_real_, nrow = length(ages), ncol = length(evalYears),
                            dimnames = list(as.character(ages), as.character(evalYears)))

# ── Loop ──────────────────────────────────────────────────────
for (i in seq_along(origins)) {
  T <- origins[i]
  
  fitCBD_f <- fit(CBD, data = mort_CBD_for, ages.fit = 25:65, years.fit = 1933:T)
  fcCBD    <- forecast(fitCBD_f, h = 1, level = 95)
  
  qx_hat_CBD[, as.character(T+1)] <- get_fc_rates_vec(fcCBD, ages)
  
  ci_cbd <- extract_cbd_ci(fitCBD_f, fcCBD, ages)
  qx_hat_CBD_lower[, as.character(T+1)] <- ci_cbd$qx_lower
  qx_hat_CBD_upper[, as.character(T+1)] <- ci_cbd$qx_upper
}

# ── Sanity check ──────────────────────────────────────────────
cat("CBD CI valid:", all(qx_hat_CBD_upper >= qx_hat_CBD_lower, na.rm = TRUE), "\n")

# Visual check for age 40, year 2010
cat("\nAge 40 CBD forecast for 2023:\n")
cat("  Lower:", qx_hat_CBD_lower["40", "2023"], "\n")
cat("  Mean: ", qx_hat_CBD["40",       "2023"], "\n")
cat("  Upper:", qx_hat_CBD_upper["40",  "2023"], "\n")
```

# Forecast Result 
```{r}
# ── Settings ──────────────────────────────────────────────────
rep_ages  <- c(30, 40, 50, 60)
evalYears <- 2003:2019

# ── Compute observed mortality rate ───────────────────────────
obs_mx <- obs_death / obs_exp

# ── Convert qx to mx ──────────────────────────────────────────
mx_hat_CBD       <- qx_hat_CBD       / (1 - qx_hat_CBD)
mx_hat_CBD_lower <- qx_hat_CBD_lower / (1 - qx_hat_CBD_lower)
mx_hat_CBD_upper <- qx_hat_CBD_upper / (1 - qx_hat_CBD_upper)

# ── Helper: extract mean series ───────────────────────────────
extract_age_series <- function(mat, age, years) {
  data.frame(
    year = years,
    rate = as.numeric(mat[as.character(age), as.character(years)])
  )
}

# ── Helper: extract CI series ─────────────────────────────────
extract_age_ci <- function(mat_lower, mat_upper, age, years) {
  data.frame(
    year  = years,
    lower = as.numeric(mat_lower[as.character(age), as.character(years)]),
    upper = as.numeric(mat_upper[as.character(age), as.character(years)])
  )
}

# ── Build long data frame for mean lines ──────────────────────
plot_data <- lapply(rep_ages, function(age) {
  bind_rows(
    extract_age_series(obs_mx, age, evalYears) %>% mutate(model = "Actual"),
    extract_age_series(mx_hat_lc_rw, age, evalYears) %>% mutate(model = "Lee-Carter"),
    extract_age_series(mx_hat_rh_ar, age, evalYears) %>% mutate(model = "Renshaw-Haberman"),
    extract_age_series(mx_hat_CBD,   age, evalYears) %>% mutate(model = "CBD")
  ) %>% mutate(age = paste0("Age ", age))
}) %>% bind_rows()

# ── Build long data frame for CI ribbons ──────────────────────
ci_data <- lapply(rep_ages, function(age) {
  bind_rows(
    extract_age_ci(mx_hat_lc_rw_lower, mx_hat_lc_rw_upper, age, evalYears) %>% 
      mutate(model = "Lee-Carter"),
    extract_age_ci(mx_hat_rh_ar_lower, mx_hat_rh_ar_upper, age, evalYears) %>%
      mutate(model = "Renshaw-Haberman"),
    extract_age_ci(mx_hat_CBD_lower,   mx_hat_CBD_upper,   age, evalYears) %>% 
      mutate(model = "CBD")
  ) %>% mutate(age = paste0("Age ", age))
}) %>% bind_rows()

# ── Plot ──────────────────────────────────────────────────────
ggplot() +
  # CI ribbons (behind lines)
  geom_ribbon(data = ci_data,
              aes(x = year, ymin = log(lower), ymax = log(upper), fill = model),
              alpha = 0.2) +
  # Mean forecast lines
  geom_line(data = plot_data,
            aes(x = year, y = log(rate), color = model, linetype = model),
            linewidth = 0.8) +
  # Actual points
  geom_point(data = filter(plot_data, model == "Actual"),
             aes(x = year, y = log(rate)),
             size = 1.2, color = "black") +
  facet_wrap(~ age, ncol = 2, scales = "free_y") +
  scale_color_manual(values = c(
    "Actual"           = "black",
    "Lee-Carter"       = "#E41A1C",
    "Renshaw-Haberman" = "#377EB8",
    "CBD"              = "#FF7F00"
  )) +
  scale_fill_manual(values = c(
    "Lee-Carter"       = "#E41A1C",
    "Renshaw-Haberman" = "#377EB8",
    "CBD"              = "#FF7F00"
  )) +
  scale_linetype_manual(values = c(
    "Actual"           = "solid",
    "Lee-Carter"       = "dashed",
    "Renshaw-Haberman" = "dotdash",
    "CBD"              = "dotted"
  )) +
  labs(
    title    = "One-Year Ahead Forecast vs Actual Mortality Rate (2003-2019)",
    x        = "Year",
    y        = "Log Mortality Rate",
    color    = "Model",
    linetype = "Model",
    fill     = "Model"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.text      = element_text(size = 12, face = "bold"),
    axis.text.x     = element_text(angle = 45, hjust = 1),
    plot.title      = element_text(size = 13, face = "bold", hjust = 0.5)
  )

# ── Save ──────────────────────────────────────────────────────
ggsave("forecast_vs_actual.png", width = 12, height = 8, dpi = 300)
```

```{r}
# ── Settings ──────────────────────────────────────────────────
rep_ages  <- c(30, 40, 50, 60)
evalYears <- 2003:2019
# ── Compute observed mortality rate ───────────────────────────
obs_mx <- obs_death / obs_exp

# ── Helper: extract one age across all years ──────────────────
extract_age_series <- function(mat, age, years) {
  data.frame(
    year = years,
    rate = as.numeric(mat[as.character(age), as.character(years)])
  )
}

# convert qx to mx before plotting
mx_hat_CBD <- qx_hat_CBD / (1 - qx_hat_CBD)

# ── Build long data frame ─────────────────────────────────────
plot_data <- lapply(rep_ages, function(age) {
  bind_rows(
    extract_age_series(obs_mx,       age, evalYears) %>% mutate(model = "Actual"),
    extract_age_series(mx_hat_lc_rw, age, evalYears) %>% mutate(model = "Lee-Carter"),
    extract_age_series(mx_hat_rh_ar, age, evalYears) %>% mutate(model = "Renshaw-Haberman"),
    extract_age_series(mx_hat_CBD,   age, evalYears) %>% mutate(model = "CBD")
  ) %>% mutate(age = paste0("Age ", age))
}) %>% bind_rows()

# ── Plot ──────────────────────────────────────────────────────
ggplot(plot_data, aes(x = year, y = log(rate), color = model, linetype = model)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = filter(plot_data, model == "Actual"), 
             size = 1.2, show.legend = FALSE) +
  facet_wrap(~ age, ncol = 2, scales = "free_y") +
  scale_color_manual(values = c(
    "Actual"           = "black",
    "Lee-Carter"       = "#E41A1C",
    "Renshaw-Haberman" = "#377EB8",
    "CBD"              = "#FF7F00"
  )) +
  scale_linetype_manual(values = c(
    "Actual"           = "solid",
    "Lee-Carter"       = "dashed",
    "Renshaw-Haberman" = "dotdash",
    "CBD"              = "dotted"
  )) +
  labs(
    title    = "One-Step Ahead Forecast vs Actual Mortality Rate (2003-2019)",
    x        = "Year",
    y        = "Log Mortality Rate",
    color    = "Model",
    linetype = "Model"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    strip.text      = element_text(size = 12, face = "bold"),
    axis.text.x     = element_text(angle = 45, hjust = 1),
    plot.title      = element_text(size = 13, face = "bold", hjust = 0.5)
  )

# ── Save ──────────────────────────────────────────────────────
ggsave("forecast_vs_actual.png", width = 12, height = 8, dpi = 300)
```

# DM Test
Poisson deviance (counts-based): 
$$
dev(x,t) = 2[D_{x,t}\text{log}(\frac{D_{x,t}}{\hat{D}_{x,t}})-(D_{x,t}-\hat{D}_{x,t})]
$$ 

### Poisson Deviance

```{r}
# Define loss function 
# Poisson Deviance
poisson_deviance_cell <- function(D, Dhat) {
  # D, Dhat numeric vectors/matrices; returns same shape
  Dhat <- pmax(Dhat, 1e-12)  # avoid division by 0
  out <- 2 * (D * ifelse(D == 0, 0, log(D / Dhat)) - (D - Dhat))
  out
}

# average
loss_per_year_poisson_dev_avg <- function(D_obs, D_hat) {
  dev_mat <- poisson_deviance_cell(D_obs, D_hat)  # age x year
  colMeans(dev_mat, na.rm = TRUE)                  # length T vector (years)
}
```

### Binomial Deviance

$$
dev(x,t)) = 2[d_{xt}\text{log}(\frac{d_{xt}}{\hat{d}_{xt}}) + (E_{x,t}^0-d_{xt})\text{log}(\frac{E_{x,t}^0-d_{xt}}{E_{x,t}^0-\hat{d_{xt}}})]
$$

```{r}
# Define loss function 
# Binomial Deviance
binom_deviance_cell <- function(D, E, qhat) {
  qhat <- pmin(pmax(qhat, 1e-12), 1 - 1e-12)
  mu <- E * qhat

  term1 <- ifelse(D == 0, 0, D * log(D / mu))
  term2 <- ifelse(D == E, 0, (E - D) * log((E - D) / (E - mu)))

  2 * (term1 + term2)
}

# average
loss_per_year_binom_dev_avg <- function(D_obs, E_obs, q_hat) {
  dev_mat <- binom_deviance_cell(D_obs, E_obs, q_hat)
  colMeans(dev_mat, na.rm = TRUE)
}
```

### Compare LC with mrwd and AR(1)

```{r}
# forecast death count 
hat_death_rw <- obs_exp*mx_hat_lc_rw
hat_death_ar <- obs_exp*mx_hat_lc_ar

# the yearly loss series for LC (avg)
# rw
lc_dev_rw <- loss_per_year_poisson_dev_avg(obs_death, hat_death_rw)
# AR
lc_dev_ar <- loss_per_year_poisson_dev_avg(obs_death, hat_death_ar)

# Put them in time series form
# rw
lc_dev_rw <- ts(lc_dev_rw, start = min(evalYears), frequency = 1)
# ar
lc_dev_ar <- ts(lc_dev_ar, start = min(evalYears), frequency = 1)

# dm test
dm.test(as.numeric(lc_dev_rw), as.numeric(lc_dev_ar), alternative = "two.sided", 
        h = 1, power = 1)
```

### Compare LC and CBD

```{r}
# the yearly loss series for CBD
# avg
L_CBD_dev_avg <- loss_per_year_binom_dev_avg(obs_death, obs_exp_CBD, qx_hat_CBD)

# Put them in time series form (optional but nice)
# avg
L_CBD_dev_avg <- ts(L_CBD_dev_avg, start = min(evalYears), frequency = 1)

# compare LC and CBD
dm.test(as.numeric(L_CBD_dev_avg), as.numeric(lc_dev_ar), alternative = "two.sided", 
        h = 1, power = 1)
```

### Compare LC and RH
```{r}
# forecast death count 
# RH
hat_death_rh <- obs_exp*mx_hat_rh_ar
# LC
hat_death_lc_rw <- obs_exp*mx_hat_lc_rw

# extract death count from 2003-2019
# RH
hat_death_rh <- hat_death_rh[,1:17]
# LC
hat_death_lc_rw <- hat_death_lc_rw[,1:17]
# obs
obs_death_0319 <- obs_death[,1:17]

# the yearly loss series for LC (avg)
# RH
rh_dev <- loss_per_year_poisson_dev_avg(obs_death_0319, hat_death_rh)
# LC
lc_dev_rw <- loss_per_year_poisson_dev_avg(obs_death_0319, hat_death_lc_rw)

# dm test
dm.test(as.numeric(rh_dev), as.numeric(lc_dev_rw), alternative = "two.sided", 
        h = 1, power = 1)
```

### Compare RH and CBD
```{r}
# the yearly loss series for CBD
# avg
CBD_dev_0319 <- loss_per_year_binom_dev_avg(obs_death[,1:17], obs_exp_CBD[,1:17], 
                                             qx_hat_CBD[,1:17])

# dm test
dm.test(as.numeric(rh_dev), as.numeric(CBD_dev_0319), alternative = "two.sided", 
        h = 1, power = 1)
```

### Small sample adjustment
```{r}
dm_hln_test <- function(e1, e2, h = 1, power = 1) {
  # e1, e2 = forecast errors from model 1 and model 2
  # h      = forecast horizon
  # power  = 1 for absolute loss
  
  T <- length(e1)
  
  # Loss differential
  d <- abs(e1)^power - abs(e2)^power
  
  # Mean loss differential
  d_bar <- mean(d)
  
  # Estimate autocovariance of d up to lag h-1
  gamma <- numeric(h)
  for (k in 0:(h - 1)) {
    gamma[k + 1] <- mean((d[(k + 1):T] - d_bar) * (d[1:(T - k)] - d_bar))
  }
  
  # Long-run variance estimate
  var_d <- (gamma[1] + 2 * sum(gamma[-1])) / T
  
  # Original DM statistic
  DM <- d_bar / sqrt(var_d)
  
  # HLN correction factor
  correction <- sqrt((T + 1 - 2*h + h*(h-1)/T) / T)
  
  # Corrected statistic
  HLN <- DM * correction
  
  # p-value from t(T-1) distribution
  p_value <- 2 * pt(-abs(HLN), df = T - 1)
  
  cat("=== Harvey-Leybourne-Newbold DM Test ===\n")
  cat("Forecast horizon (h)   :", h, "\n")
  cat("Sample size (T)        :", T, "\n")
  cat("Original DM statistic  :", round(DM, 4), "\n")
  cat("HLN correction factor  :", round(correction, 4), "\n")
  cat("HLN statistic          :", round(HLN, 4), "\n")
  cat("Degrees of freedom     :", T - 1, "\n")
  cat("p-value                :", round(p_value, 4), "\n")
  
  return(invisible(list(
    DM        = DM,
    HLN       = HLN,
    p_value   = p_value,
    df        = T - 1,
    correction = correction
  )))
}
```

```{r}
dm_hln_test(as.numeric(rh_dev), as.numeric(CBD_dev_0319), h = 1, power = 1)
```

# Combined Model
```{r}
# load opioid related mortality rate
logmxt_opioid <- as.matrix(read.csv("national_opioid_logmxt.csv", 
                                    row.names = 1, check.names = FALSE))
opioid_mort <- exp(logmxt_opioid)
```

## LC
```{r}
# setting
ages      <- 25:65
fit_years <- 1933:2023
out_years <- 2003:2020

# Prepare dataX
mort_F <- StMoMoData(US_data, type = "central", series = "total")

# Fit LC model 
LC <- lc(link = "log")
LCfit <- fit(LC, data = mort_F, ages.fit = ages, years.fit = fit_years)

# Extract Fitted Rate
fitted_rates <- fitted(LCfit, type = "rates")

# Subset to ages 25-65 and years 2003-2020
mx_base_lc <- fitted_rates[as.character(ages), as.character(out_years)]

# combine two mortality
mxt_lc <- mx_base_lc + opioid_mort

# ratio
ratio_lc <- opioid_mort / mxt_lc
write.csv(ratio_lc, file = "../hmd/ratio_adjust_mortality_lc.csv", 
          row.names = TRUE, quote = FALSE)
```

```{r}
# Convert matrix to long format
ratio_lc_long <- melt(ratio_lc,
                      varnames   = c("Age", "Year"),
                      value.name = "Ratio")

# Plot heatmap
ggplot(ratio_lc_long, aes(x = as.factor(Year), y = as.factor(Age), fill = Ratio)) +
  geom_tile() +
  scale_fill_gradient(
    low    = "white",
    high   = "red",
    limits = c(0, 0.25),
    breaks = seq(0, 0.25, by = 0.05),
    labels = scales::percent_format(accuracy = 1),
    name   = "Opioid\nShare"
  ) +
  labs(
    title = "Opioid Mortality as Share of Total LC Mortality Rate (2003-2020)",
    x     = "Year",
    y     = "Age"
  ) +
  theme_bw() +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y     = element_text(size = 8),
    plot.title      = element_text(size = 13, face = "bold", hjust = 0.5),
    legend.position = "right"
  )

ggsave("../hmd/ratio_lc_heatmap.png", width = 10, height = 8, dpi = 300)
```

## RH
```{r}
# Settings
ages <- 25:65
fit_years <- 1933:2020
out_years <- 2003:2020
max_try <- 15

# Prepare data
mort_rh <- StMoMoData(US_data, type = "central", series = "total")
RHnp_mod <- rh(link = "log", cohortAgeFun = "NP")

# Fit RH model multiple times, keep best
best_loglik <- -Inf
best_fit <- NULL

for (j in seq_len(max_try)) {
  cat("\n--- Attempt", j, "of", max_try, "---\n")
  
  fitRH_j <- tryCatch(
    fit(RHnp_mod, data = mort_rh,
        ages.fit  = ages,
        years.fit = fit_years,
        control   = list(maxit = 200)),
    error = function(e) NULL
  )
  
  # skip if fit failed or did not converge
  if (is.null(fitRH_j) || isTRUE(fitRH_j$fail)) {
    cat("Fit failed\n")
    next
  }
  if (!isTRUE(fitRH_j$conv)) {
    cat("Did not converge\n")
    next
  }
  
  cat("Converged! loglik:", round(fitRH_j$loglik, 2), "\n")
  
  if (fitRH_j$loglik > best_loglik) {
    best_loglik <- fitRH_j$loglik
    best_fit    <- fitRH_j
    cat("*** New best loglik:", round(best_loglik, 2), "***\n")
  }
}

cat("\n==========================================\n")
cat("Best loglik:", round(best_loglik, 2), "\n")
cat("RH converged:", best_fit$conv, "\n")
cat("==========================================\n")


# Extract fitted rates
fitted_rates_rh <- fitted(best_fit, type = "rates")

# Subset to ages 25-65 and years 2003-2020
mx_base_rh <- fitted_rates_rh[as.character(ages), as.character(out_years)]

# Combine and compute ratio
mxt_rh    <- mx_base_rh + exp(logmxt_opioid)
ratio_rh  <- exp(logmxt_opioid) / mxt_rh

# Save ratio matrix
write.csv(ratio_rh, file = "../hmd/ratio_adjust_mortality_rh.csv",
          row.names = TRUE, quote = FALSE)
```

```{r}
# Convert matrix to long format
ratio_rh_long<- melt(ratio_rh,
                      varnames   = c("Age", "Year"),
                      value.name = "Ratio")

# Plot heatmap
ggplot(ratio_rh_long, aes(x = as.factor(Year), y = as.factor(Age), fill = Ratio)) +
  geom_tile() +
  scale_fill_gradient(
    low    = "white",
    high   = "red",
    limits = c(0, 0.2),
    breaks = seq(0, 0.2, by = 0.05),
    labels = scales::percent_format(accuracy = 1),
    name   = "Opioid\nShare"
  ) +
  labs(
    title = "Opioid Mortality as Share of Total RH Mortality Rate (2003-2020)",
    x     = "Year",
    y     = "Age"
  ) +
  theme_bw() +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y     = element_text(size = 8),
    plot.title      = element_text(size = 13, face = "bold", hjust = 0.5),
    legend.position = "right"
  )

ggsave("../hmd/ratio_rh_heatmap.png", width = 10, height = 8, dpi = 300)
```

## CBD
```{r}
# Settings
ages      <- 25:65
fit_years <- 1933:2020
out_years <- 2003:2020

# Prepare data (CBD uses initial exposure
mort_cbd_for <- StMoMoData(US_data, type = "initial", series = "total")

# Scale exposures
N <- 1e6
mort_cbd_for$Dxt <- round(mort_cbd_for$Dxt * N)
mort_cbd_for$Ext <- round(mort_cbd_for$Ext * N)

# Fit CBD model
CBD    <- cbd(link = "logit")
CBDfit <- fit(CBD, data = mort_cbd_for,
              ages.fit  = ages,
              years.fit = fit_years)
cat("CBD converged:", CBDfit$conv, "\n")

# Extract fitted rates
# CBD fitted() returns probability (qx) scale
fitted_qx_cbd <- fitted(CBDfit, type = "rates")

# convert qx to mx
fitted_rates_cbd <- fitted_qx_cbd / (1 - fitted_qx_cbd)

# ubset to ages 25-65 and years 2003-2020
mx_base_cbd <- fitted_rates_cbd[as.character(ages), as.character(out_years)]

# Combine and compute ratio
mxt_cbd   <- mx_base_cbd + exp(logmxt_opioid)
ratio_cbd <- exp(logmxt_opioid) / mxt_cbd

# save ratio matrix
write.csv(ratio_cbd, file = "../hmd/ratio_adjust_mortality_cbd.csv",
          row.names = TRUE, quote = FALSE)
```

```{r}
# Convert matrix to long format
ratio_cbd_long <- melt(ratio_cbd,
                       varnames   = c("Age", "Year"),
                       value.name = "Ratio")

# Plot heatmap
ggplot(ratio_cbd_long, aes(x = as.factor(Year), y = as.factor(Age), fill = Ratio)) +
  geom_tile() +
  scale_fill_gradient(
    low    = "white",
    high   = "red",
    limits = c(0, 0.25),
    breaks = seq(0, 0.25, by = 0.05),
    labels = scales::percent_format(accuracy = 1),
    name   = "Opioid\nShare"
  ) +
  labs(
    title = "Opioid Mortality as Share of Total CBD Mortality Rate (2003-2020)",
    x     = "Year",
    y     = "Age"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    legend.position = "right"
  )

ggsave("../hmd/ratio_cbd_heatmap.png", width = 10, height = 8, dpi = 300)
```
