---
title: "Combine Data to HMD Format"
author: "PSTAT 296A"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, message=FALSE, warning=FALSE}
rm(list=ls());gc()
library(tidyverse)
library(demography)
library(StMoMo)
```

# Load Data

```{r load-data}
# Deaths from WONDER (monthly, by region/sex/age group)
wonder <- read.csv("../Data/missing_data/Inferred-WONDER-Census Region-Sex-5Yr.csv")

# Population by region, age, sex
pop <- read.csv("../Data/covariate/population_by_region_age_sex.csv")

# State-level total population (for regional total population)
state_pop <- read.csv("../Data/covariate/state_population.csv")

# Covariates by region and year
covariates <- read.csv("../Data/covariate/covariate_combine_region_pre_whitening.csv")
```

# Data Cleaning and Harmonization

The three data sources use different naming conventions for Census Regions:

- WONDER: `"Census Region 1: Northeast"`, `"Census Region 2: Midwest"`, etc.
- Population: `"North Central"` (= Midwest), `"Northeast"`, `"South"`, `"West"`
- Covariates: `"Midwest"`, `"Northeast"`, `"South"`, `"West"`

We standardize all to: **Northeast, Midwest, South, West**.

```{r clean-data}
# Define target age groups (5-year bins, ages 15-54)
age_groups <- c("15-19", "20-24", "25-29", "30-34",
                "35-39", "40-44", "45-49", "50-54")
age_starts <- c(15, 20, 25, 30, 35, 40, 45, 50)
age_map    <- setNames(age_starts, age_groups)

target_years <- 2005:2019

# ---- Clean WONDER deaths (keep monthly) ----
wonder_monthly <- wonder %>%
  mutate(
    Region = case_when(
      Census.Region == "Census Region 1: Northeast" ~ "Northeast",
      Census.Region == "Census Region 2: Midwest"   ~ "Midwest",
      Census.Region == "Census Region 3: South"     ~ "South",
      Census.Region == "Census Region 4: West"      ~ "West"
    ),
    Deaths = suppressWarnings(as.numeric(Deaths)),
    Month  = as.integer(sub(".*/", "", Month.Code))
  ) %>%
  filter(
    Year %in% target_years,
    Age.Group.Code %in% age_groups,
    Sex.Code %in% c("M", "F")
  ) %>%
  mutate(
    Age = age_map[Age.Group.Code],
    Sex = ifelse(Sex.Code == "M", "Male", "Female")
  ) %>%
  select(Region, Sex, Age, Age.Group.Code, Year, Month, Deaths)

# ---- Aggregate to yearly ----
wonder_yearly <- wonder_monthly %>%
  group_by(Region, Sex, Age, Age.Group.Code, Year) %>%
  summarise(
    Months_Observed = sum(!is.na(Deaths)),
    Deaths = sum(Deaths, na.rm = TRUE),
    .groups = "drop"
  )

# ---- Clean Population ----
pop_clean <- pop %>%
  mutate(
    Region = ifelse(census_region == "North Central", "Midwest", census_region)
  ) %>%
  filter(
    year %in% target_years,
    age5 %in% age_groups,
    sex %in% c("Male", "Female")
  ) %>%
  mutate(Age = age_map[age5]) %>%
  rename(Year = year, Sex = sex, Population = population)

# ---- Regional total population ----
state_region_map <- tibble(
  state_name = c(
    "Connecticut", "Maine", "Massachusetts", "New Hampshire", "New Jersey",
    "New York", "Pennsylvania", "Rhode Island", "Vermont",
    "Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota",
    "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin",
    "Alabama", "Arkansas", "Delaware", "District of Columbia", "Florida",
    "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi",
    "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas",
    "Virginia", "West Virginia",
    "Alaska", "Arizona", "California", "Colorado", "Hawaii", "Idaho",
    "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming"
  ),
  Region = c(
    rep("Northeast", 9), rep("Midwest", 12),
    rep("South", 17),    rep("West", 13)
  )
)

regional_pop <- state_pop %>%
  filter(year %in% target_years) %>%
  left_join(state_region_map, by = "state_name") %>%
  group_by(Region, year) %>%
  summarise(Total_Population = sum(population), .groups = "drop") %>%
  rename(Year = year)

# ---- Check for suppressed months ----
suppressed_summary <- wonder_yearly %>% filter(Months_Observed < 12)
if (nrow(suppressed_summary) > 0) {
  cat("WARNING: Some combinations had months with suppressed data:\n")
  suppressed_summary %>%
    group_by(Region, Sex, Age.Group.Code) %>%
    summarise(
      n_years_affected = n(),
      avg_months_observed = mean(Months_Observed),
      .groups = "drop"
    ) %>%
    print(n = 50)
} else {
  cat("All combinations have 12 months of data. No suppression issues.\n")
}
```

# Clean Covariates

```{r clean-covariates}
cov_clean <- covariates %>%
  rename(Region = Census_Region) %>%
  filter(Year %in% target_years) %>%
  select(Region, Year,
         rate_health_care_coverage_pw,
         rate_alochol_consum_pw,
         socio_score)

cat("Covariate columns:", paste(names(cov_clean), collapse = ", "), "\n")
```

# Build Flat Data Frames

Yearly and monthly versions for verification.

```{r build-yearly}
flat_yearly <- wonder_yearly %>%
  left_join(
    pop_clean %>% select(Region, Sex, Age, Year, Population),
    by = c("Region", "Sex", "Age", "Year")
  ) %>%
  left_join(regional_pop, by = c("Region", "Year")) %>%
  mutate(Death_Rate = Deaths / Population) %>%
  select(Region, Sex, Age_Group = Age.Group.Code, Age, Year,
         Deaths, Months_Observed, Population, Total_Population,
         Death_Rate)

cat(sprintf("Yearly: %d rows x %d cols\n",
            nrow(flat_yearly), ncol(flat_yearly)))
```

```{r build-monthly}
flat_monthly <- wonder_monthly %>%
  left_join(
    pop_clean %>% select(Region, Sex, Age, Year, Population),
    by = c("Region", "Sex", "Age", "Year")
  ) %>%
  left_join(regional_pop, by = c("Region", "Year")) %>%
  select(Region, Sex, Age_Group = Age.Group.Code, Age, Year, Month,
         Deaths, Population, Total_Population)

cat(sprintf("Monthly: %d rows x %d cols\n",
            nrow(flat_monthly), ncol(flat_monthly)))
```

# Build StMoMo Objects — Yearly

For each **Region x Sex** combination, we create death and exposure
matrices compatible with `demography` and `StMoMo` packages.

```{r yearly-matrices}
regions <- c("Northeast", "Midwest", "South", "West")
sexes   <- c("Male", "Female")

make_matrices_yearly <- function(deaths_df, pop_df, region, sex) {
  Dxt <- deaths_df %>%
    filter(Region == region, Sex == sex) %>%
    select(Age, Year, Deaths) %>%
    pivot_wider(names_from = Year, values_from = Deaths) %>%
    arrange(Age) %>%
    column_to_rownames("Age") %>%
    as.matrix()

  Ext <- pop_df %>%
    filter(Region == region, Sex == sex) %>%
    select(Age, Year, Population) %>%
    pivot_wider(names_from = Year, values_from = Population) %>%
    arrange(Age) %>%
    column_to_rownames("Age") %>%
    as.matrix()

  stopifnot(all(dim(Dxt) == dim(Ext)))
  list(Dxt = Dxt, Ext = Ext)
}

hmd_matrices_yr <- list()
for (reg in regions) {
  for (sex in sexes) {
    key <- paste(reg, sex, sep = "_")
    hmd_matrices_yr[[key]] <- make_matrices_yearly(
      wonder_yearly, pop_clean, reg, sex)
    cat(sprintf("  Yearly %s: [%d ages x %d years]\n", key,
                nrow(hmd_matrices_yr[[key]]$Dxt),
                ncol(hmd_matrices_yr[[key]]$Dxt)))
  }
}
```

```{r yearly-demogdata}
hmd_region_yr <- list()

for (reg in regions) {
  mat_m <- hmd_matrices_yr[[paste(reg, "Male",   sep = "_")]]
  mat_f <- hmd_matrices_yr[[paste(reg, "Female", sep = "_")]]

  hmd_region_yr[[reg]] <- demogdata(
    data  = mat_m$Dxt / mat_m$Ext,
    pop   = mat_m$Ext,
    ages  = age_starts,
    years = target_years,
    type  = "mortality",
    label = reg,
    name  = "male"
  )
  hmd_region_yr[[reg]]$rate$female <- mat_f$Dxt / mat_f$Ext
  hmd_region_yr[[reg]]$pop$female  <- mat_f$Ext
}
```

```{r yearly-stmomodata}
# Without covariates
stmomo_yr <- list()
for (reg in regions) {
  for (series in c("male", "female")) {
    key <- paste(reg, series, sep = "_")
    stmomo_yr[[key]] <- StMoMoData(
      hmd_region_yr[[reg]], series = series, type = "central")
  }
}

# With covariates
stmomo_yr_cov <- list()
for (reg in regions) {
  cov_reg <- cov_clean %>%
    filter(Region == reg) %>%
    arrange(Year) %>%
    select(-Region, -Year)
  for (series in c("male", "female")) {
    key <- paste(reg, series, sep = "_")
    stmomo_yr_cov[[key]] <- stmomo_yr[[key]]
    stmomo_yr_cov[[key]]$covariates <- cov_reg
  }
}

cat("Yearly StMoMoData created for:",
    paste(names(stmomo_yr), collapse = ", "), "\n")
```

# Build StMoMo Objects — Monthly

Monthly matrices use fractional years (e.g. 2005.000, 2005.083, ...).
Monthly exposure = annual population / 12.

```{r monthly-matrices}
# Time points as fractional years
monthly_times <- sort(unique(
  wonder_monthly$Year + (wonder_monthly$Month - 1) / 12
))

make_matrices_monthly <- function(deaths_df, pop_df, region, sex) {
  df <- deaths_df %>%
    filter(Region == region, Sex == sex) %>%
    mutate(Time = Year + (Month - 1) / 12)

  Dxt <- df %>%
    select(Age, Time, Deaths) %>%
    pivot_wider(names_from = Time, values_from = Deaths) %>%
    arrange(Age) %>%
    column_to_rownames("Age") %>%
    as.matrix()

  # Monthly exposure = annual population / 12
  pop_reg <- pop_df %>%
    filter(Region == region, Sex == sex)

  Ext_yr <- pop_reg %>%
    select(Age, Year, Population) %>%
    pivot_wider(names_from = Year, values_from = Population) %>%
    arrange(Age) %>%
    column_to_rownames("Age") %>%
    as.matrix()

  # Repeat each year's population across 12 months, then / 12
  Ext <- Ext_yr[, rep(seq_len(ncol(Ext_yr)), each = 12)] / 12
  colnames(Ext) <- colnames(Dxt)

  stopifnot(all(dim(Dxt) == dim(Ext)))
  list(Dxt = Dxt, Ext = Ext)
}

hmd_matrices_mo <- list()
for (reg in regions) {
  for (sex in sexes) {
    key <- paste(reg, sex, sep = "_")
    hmd_matrices_mo[[key]] <- make_matrices_monthly(
      wonder_monthly, pop_clean, reg, sex)
    cat(sprintf("  Monthly %s: [%d ages x %d months]\n", key,
                nrow(hmd_matrices_mo[[key]]$Dxt),
                ncol(hmd_matrices_mo[[key]]$Dxt)))
  }
}
```

```{r monthly-demogdata}
hmd_region_mo <- list()

for (reg in regions) {
  mat_m <- hmd_matrices_mo[[paste(reg, "Male",   sep = "_")]]
  mat_f <- hmd_matrices_mo[[paste(reg, "Female", sep = "_")]]

  hmd_region_mo[[reg]] <- demogdata(
    data  = mat_m$Dxt / mat_m$Ext,
    pop   = mat_m$Ext,
    ages  = age_starts,
    years = monthly_times,
    type  = "mortality",
    label = reg,
    name  = "male"
  )
  hmd_region_mo[[reg]]$rate$female <- mat_f$Dxt / mat_f$Ext
  hmd_region_mo[[reg]]$pop$female  <- mat_f$Ext
}
```

```{r monthly-stmomodata}
# Without covariates
stmomo_mo <- list()
for (reg in regions) {
  for (series in c("male", "female")) {
    key <- paste(reg, series, sep = "_")
    stmomo_mo[[key]] <- StMoMoData(
      hmd_region_mo[[reg]], series = series, type = "central")
  }
}

# With covariates (annual covariates repeated for each month)
cov_monthly <- cov_clean %>%
  slice(rep(seq_len(n()), each = 12)) %>%
  select(-Region, -Year)

stmomo_mo_cov <- list()
for (reg in regions) {
  cov_reg_mo <- cov_clean %>%
    filter(Region == reg) %>%
    arrange(Year) %>%
    slice(rep(seq_len(n()), each = 12)) %>%
    select(-Region, -Year)
  for (series in c("male", "female")) {
    key <- paste(reg, series, sep = "_")
    stmomo_mo_cov[[key]] <- stmomo_mo[[key]]
    stmomo_mo_cov[[key]]$covariates <- cov_reg_mo
  }
}

cat("Monthly StMoMoData created for:",
    paste(names(stmomo_mo), collapse = ", "), "\n")
```

# Verification

```{r summary-table}
flat_yearly %>%
  group_by(Region, Sex) %>%
  summarise(
    Total_Deaths = sum(Deaths),
    Mean_Rate    = mean(Death_Rate),
    Min_Rate     = min(Death_Rate),
    Max_Rate     = max(Death_Rate),
    .groups = "drop"
  ) %>%
  knitr::kable(digits = 6,
               caption = "Summary by Region and Sex (2005-2019, Ages 15-54)")
```

```{r plot-verify, fig.width=10, fig.height=8}
flat_yearly %>%
  ggplot(aes(x = Year, y = log(Death_Rate),
             color = Age_Group, linetype = Sex)) +
  geom_line() +
  facet_wrap(~Region, ncol = 2) +
  labs(
    title = "Log Mortality Rate by Region, Age Group, and Sex (2005-2019)",
    y     = "Log(Death Rate)",
    color = "Age Group"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Save Outputs

```{r save}
out <- "../Data/wonder_opioid"

# Yearly
saveRDS(stmomo_yr, file.path(out, "hmd_yearly_no_cov.rds"))
saveRDS(stmomo_yr_cov, file.path(out, "hmd_yearly_with_cov.rds"))
saveRDS(hmd_region_yr, file.path(out, "demogdata_yearly.rds"))
saveRDS(flat_yearly, file.path(out, "flat_yearly.rds"))

# Monthly
saveRDS(stmomo_mo, file.path(out, "hmd_monthly_no_cov.rds"))
saveRDS(stmomo_mo_cov, file.path(out, "hmd_monthly_with_cov.rds"))
saveRDS(hmd_region_mo, file.path(out, "demogdata_monthly.rds"))
saveRDS(flat_monthly, file.path(out, "flat_monthly.rds"))

cat("Saved:\n")
cat("  Yearly:\n")
cat("    hmd_yearly_no_cov.rds   - StMoMoData [8 ages x 15 years]\n")
cat("    hmd_yearly_with_cov.rds - StMoMoData + covariates\n")
cat("    demogdata_yearly.rds    - demogdata objects\n")
cat("    flat_yearly.rds         - flat data frame\n")
cat("  Monthly:\n")
cat("    hmd_monthly_no_cov.rds   - StMoMoData [8 ages x 180 months]\n")
cat("    hmd_monthly_with_cov.rds - StMoMoData + covariates\n")
cat("    demogdata_monthly.rds    - demogdata objects\n")
cat("    flat_monthly.rds         - flat data frame\n")
cat("\nCovariate columns:\n")
print(names(stmomo_yr_cov[["Northeast_male"]]$covariates))
```
