---
title: "Combine Data to HMD Format"
author: "PSTAT 296A"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, message=FALSE, warning=FALSE}
rm(list=ls());gc()
library(tidyverse)
library(demography)
library(StMoMo)
```

# Load Data

```{r load-data}
# Opioid-related deaths from WONDER (monthly, by region/sex/age group)
wonder <- read.csv("../Data/missing_data/Inferred-WONDER-Census Region-Sex-5Yr.csv")

# All-cause deaths from WONDER (monthly, by region/sex/age group)
allcause <- read.csv("../Data/wonder_opioid/all-cause-region-yr-sex.csv")

# Population by region, age, sex
pop <- read.csv("../Data/covariate/population_by_region_age_sex.csv")

# State-level total population (for regional total population)
state_pop <- read.csv("../Data/covariate/state_population.csv")

# Covariates by region and year
covariates <- read.csv("../Data/covariate/covariate_combine_region_pre_whitening.csv")
```

# Data Cleaning and Harmonization

The three data sources use different naming conventions for Census Regions:

- WONDER: `"Census Region 1: Northeast"`, `"Census Region 2: Midwest"`, etc.
- Population: `"North Central"` (= Midwest), `"Northeast"`, `"South"`, `"West"`
- Covariates: `"Midwest"`, `"Northeast"`, `"South"`, `"West"`

We standardize all to: **Northeast, Midwest, South, West**.

```{r clean-data}
# Define target age groups (5-year bins, ages 15-54)
age_groups <- c("15-19", "20-24", "25-29", "30-34",
                "35-39", "40-44", "45-49", "50-54")
age_starts <- c(15, 20, 25, 30, 35, 40, 45, 50)
age_map    <- setNames(age_starts, age_groups)

target_years <- 2005:2019

# ---- Clean WONDER deaths (keep monthly) ----
wonder_monthly <- wonder %>%
  mutate(
    Region = case_when(
      Census.Region == "Census Region 1: Northeast" ~ "Northeast",
      Census.Region == "Census Region 2: Midwest"   ~ "Midwest",
      Census.Region == "Census Region 3: South"     ~ "South",
      Census.Region == "Census Region 4: West"      ~ "West"
    ),
    Deaths = suppressWarnings(as.numeric(Deaths)),
    Month  = as.integer(sub(".*/", "", Month.Code))
  ) %>%
  filter(
    Year %in% target_years,
    Age.Group.Code %in% age_groups,
    Sex.Code %in% c("M", "F")
  ) %>%
  mutate(
    Age = age_map[Age.Group.Code],
    Sex = ifelse(Sex.Code == "M", "Male", "Female")
  ) %>%
  select(Region, Sex, Age, Age.Group.Code, Year, Month, Deaths)

# ---- Clean all-cause deaths (keep monthly) ----
allcause_monthly <- allcause %>%
  mutate(
    Region = case_when(
      Census.Region == "Census Region 1: Northeast" ~ "Northeast",
      Census.Region == "Census Region 2: Midwest"   ~ "Midwest",
      Census.Region == "Census Region 3: South"     ~ "South",
      Census.Region == "Census Region 4: West"      ~ "West"
    ),
    Deaths = suppressWarnings(as.numeric(Deaths)),
    Month  = as.integer(sub(".*/", "", Month.Code)),
    Year   = as.integer(Year)
  ) %>%
  filter(
    Year %in% target_years,
    Five.Year.Age.Groups.Code %in% age_groups,
    Sex.Code %in% c("M", "F")
  ) %>%
  mutate(
    Age = age_map[Five.Year.Age.Groups.Code],
    Sex = ifelse(Sex.Code == "M", "Male", "Female")
  ) %>%
  select(Region, Sex, Age, Age.Group.Code = Five.Year.Age.Groups.Code,
         Year, Month, Deaths)

# ---- Compute non-opioid deaths (monthly) ----
nonopioid_monthly <- allcause_monthly %>%
  rename(AllCause_Deaths = Deaths) %>%
  left_join(
    wonder_monthly %>% rename(Opioid_Deaths = Deaths),
    by = c("Region", "Sex", "Age", "Age.Group.Code", "Year", "Month")
  ) %>%
  mutate(
    Opioid_Deaths = replace_na(Opioid_Deaths, 0),
    Deaths = AllCause_Deaths - Opioid_Deaths
  ) %>%
  select(Region, Sex, Age, Age.Group.Code, Year, Month, Deaths)

cat(sprintf("Non-opioid monthly: %d rows, min deaths = %d\n",
            nrow(nonopioid_monthly), min(nonopioid_monthly$Deaths, na.rm = TRUE)))

# ---- Aggregate opioid to yearly ----
wonder_yearly <- wonder_monthly %>%
  group_by(Region, Sex, Age, Age.Group.Code, Year) %>%
  summarise(
    Months_Observed = sum(!is.na(Deaths)),
    Deaths = sum(Deaths, na.rm = TRUE),
    .groups = "drop"
  )

# ---- Aggregate non-opioid to yearly ----
nonopioid_yearly <- nonopioid_monthly %>%
  group_by(Region, Sex, Age, Age.Group.Code, Year) %>%
  summarise(
    Months_Observed = sum(!is.na(Deaths)),
    Deaths = sum(Deaths, na.rm = TRUE),
    .groups = "drop"
  )

# ---- Clean Population ----
pop_clean <- pop %>%
  mutate(
    Region = ifelse(census_region == "North Central", "Midwest", census_region)
  ) %>%
  filter(
    year %in% target_years,
    age5 %in% age_groups,
    sex %in% c("Male", "Female")
  ) %>%
  mutate(Age = age_map[age5]) %>%
  rename(Year = year, Sex = sex, Population = population)

# ---- Regional total population ----
state_region_map <- tibble(
  state_name = c(
    "Connecticut", "Maine", "Massachusetts", "New Hampshire", "New Jersey",
    "New York", "Pennsylvania", "Rhode Island", "Vermont",
    "Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota",
    "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin",
    "Alabama", "Arkansas", "Delaware", "District of Columbia", "Florida",
    "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi",
    "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas",
    "Virginia", "West Virginia",
    "Alaska", "Arizona", "California", "Colorado", "Hawaii", "Idaho",
    "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming"
  ),
  Region = c(
    rep("Northeast", 9), rep("Midwest", 12),
    rep("South", 17),    rep("West", 13)
  )
)

regional_pop <- state_pop %>%
  filter(year %in% target_years) %>%
  left_join(state_region_map, by = "state_name") %>%
  group_by(Region, year) %>%
  summarise(Total_Population = sum(population), .groups = "drop") %>%
  rename(Year = year)

# ---- Check for suppressed months ----
suppressed_summary <- wonder_yearly %>% filter(Months_Observed < 12)
if (nrow(suppressed_summary) > 0) {
  cat("WARNING: Some combinations had months with suppressed data:\n")
  suppressed_summary %>%
    group_by(Region, Sex, Age.Group.Code) %>%
    summarise(
      n_years_affected = n(),
      avg_months_observed = mean(Months_Observed),
      .groups = "drop"
    ) %>%
    print(n = 50)
} else {
  cat("All combinations have 12 months of data. No suppression issues.\n")
}
```

# Clean Covariates

```{r clean-covariates}
cov_clean <- covariates %>%
  rename(Region = Census_Region) %>%
  filter(Year %in% target_years) %>%
  select(Region, Year,
         rate_health_care_coverage_pw,
         rate_alochol_consum_pw,
         socio_score)

cat("Covariate columns:", paste(names(cov_clean), collapse = ", "), "\n")
```

# Build Flat Data Frames

Yearly and monthly versions for verification.

```{r build-yearly}
# Helper to build flat data frames
build_flat_yearly <- function(deaths_df, pop_df, reg_pop) {
  deaths_df %>%
    left_join(pop_df %>% select(Region, Sex, Age, Year, Population),
              by = c("Region", "Sex", "Age", "Year")) %>%
    left_join(reg_pop, by = c("Region", "Year")) %>%
    mutate(Death_Rate = Deaths / Population) %>%
    select(Region, Sex, Age_Group = Age.Group.Code, Age, Year,
           Deaths, Months_Observed, Population, Total_Population, Death_Rate)
}

build_flat_monthly <- function(deaths_df, pop_df, reg_pop) {
  deaths_df %>%
    left_join(pop_df %>% select(Region, Sex, Age, Year, Population),
              by = c("Region", "Sex", "Age", "Year")) %>%
    left_join(reg_pop, by = c("Region", "Year")) %>%
    select(Region, Sex, Age_Group = Age.Group.Code, Age, Year, Month,
           Deaths, Population, Total_Population)
}

# Opioid
flat_yearly <- build_flat_yearly(wonder_yearly, pop_clean, regional_pop)
flat_monthly <- build_flat_monthly(wonder_monthly, pop_clean, regional_pop)

# Non-opioid
flat_yearly_nonopioid <- build_flat_yearly(nonopioid_yearly, pop_clean, regional_pop)
flat_monthly_nonopioid <- build_flat_monthly(nonopioid_monthly, pop_clean, regional_pop)

cat(sprintf("Opioid     — Yearly: %d rows, Monthly: %d rows\n",
            nrow(flat_yearly), nrow(flat_monthly)))
cat(sprintf("Non-opioid — Yearly: %d rows, Monthly: %d rows\n",
            nrow(flat_yearly_nonopioid), nrow(flat_monthly_nonopioid)))
```

# Build StMoMo Objects — Yearly

For each **Region x Sex** combination, we create death and exposure
matrices compatible with `demography` and `StMoMo` packages.

```{r yearly-matrices}
regions <- c("Northeast", "Midwest", "South", "West")
sexes   <- c("Male", "Female")

make_matrices_yearly <- function(deaths_df, pop_df, region, sex) {
  Dxt <- deaths_df %>%
    filter(Region == region, Sex == sex) %>%
    select(Age, Year, Deaths) %>%
    pivot_wider(names_from = Year, values_from = Deaths) %>%
    arrange(Age) %>%
    column_to_rownames("Age") %>%
    as.matrix()

  Ext <- pop_df %>%
    filter(Region == region, Sex == sex) %>%
    select(Age, Year, Population) %>%
    pivot_wider(names_from = Year, values_from = Population) %>%
    arrange(Age) %>%
    column_to_rownames("Age") %>%
    as.matrix()

  stopifnot(all(dim(Dxt) == dim(Ext)))
  list(Dxt = Dxt, Ext = Ext)
}

# Opioid
hmd_matrices_yr <- list()
for (reg in regions) {
  for (sex in sexes) {
    key <- paste(reg, sex, sep = "_")
    hmd_matrices_yr[[key]] <- make_matrices_yearly(
      wonder_yearly, pop_clean, reg, sex)
  }
}

# Non-opioid
hmd_matrices_yr_no <- list()
for (reg in regions) {
  for (sex in sexes) {
    key <- paste(reg, sex, sep = "_")
    hmd_matrices_yr_no[[key]] <- make_matrices_yearly(
      nonopioid_yearly, pop_clean, reg, sex)
  }
}

cat("Yearly matrices built (opioid + non-opioid)\n")
```

```{r yearly-demogdata}
build_demogdata_yearly <- function(matrices, age_starts, target_years, label_suffix = "") {
  result <- list()
  for (reg in regions) {
    mat_m <- matrices[[paste(reg, "Male",   sep = "_")]]
    mat_f <- matrices[[paste(reg, "Female", sep = "_")]]
    lbl <- if (label_suffix == "") reg else paste(reg, label_suffix)
    result[[reg]] <- demogdata(
      data  = mat_m$Dxt / mat_m$Ext,
      pop   = mat_m$Ext,
      ages  = age_starts,
      years = target_years,
      type  = "mortality",
      label = lbl,
      name  = "male"
    )
    result[[reg]]$rate$female <- mat_f$Dxt / mat_f$Ext
    result[[reg]]$pop$female  <- mat_f$Ext
  }
  result
}

hmd_region_yr    <- build_demogdata_yearly(hmd_matrices_yr,    age_starts, target_years)
hmd_region_yr_no <- build_demogdata_yearly(hmd_matrices_yr_no, age_starts, target_years, "non-opioid")
```

```{r yearly-stmomodata}
build_stmomo_yearly <- function(demog_list, cov_df = NULL) {
  stmomo_list <- list()
  for (reg in regions) {
    for (series in c("male", "female")) {
      key <- paste(reg, series, sep = "_")
      stmomo_list[[key]] <- StMoMoData(demog_list[[reg]], series = series, type = "central")
      if (!is.null(cov_df)) {
        stmomo_list[[key]]$covariates <- cov_df %>%
          filter(Region == reg) %>% arrange(Year) %>% select(-Region, -Year)
      }
    }
  }
  stmomo_list
}

# Opioid
stmomo_yr     <- build_stmomo_yearly(hmd_region_yr)
stmomo_yr_cov <- build_stmomo_yearly(hmd_region_yr, cov_clean)

# Non-opioid
stmomo_yr_no     <- build_stmomo_yearly(hmd_region_yr_no)
stmomo_yr_no_cov <- build_stmomo_yearly(hmd_region_yr_no, cov_clean)

cat("Yearly StMoMoData created (opioid + non-opioid)\n")
```

# Build StMoMo Objects — Monthly

Monthly matrices use fractional years (e.g. 2005.000, 2005.083, ...).
Monthly exposure = annual population / 12.

```{r monthly-matrices}
# Time points as fractional years
monthly_times <- sort(unique(
  wonder_monthly$Year + (wonder_monthly$Month - 1) / 12
))

make_matrices_monthly <- function(deaths_df, pop_df, region, sex) {
  df <- deaths_df %>%
    filter(Region == region, Sex == sex) %>%
    mutate(Time = Year + (Month - 1) / 12)

  Dxt <- df %>%
    select(Age, Time, Deaths) %>%
    pivot_wider(names_from = Time, values_from = Deaths) %>%
    arrange(Age) %>%
    column_to_rownames("Age") %>%
    as.matrix()

  # Monthly exposure = annual population / 12
  pop_reg <- pop_df %>%
    filter(Region == region, Sex == sex)

  Ext_yr <- pop_reg %>%
    select(Age, Year, Population) %>%
    pivot_wider(names_from = Year, values_from = Population) %>%
    arrange(Age) %>%
    column_to_rownames("Age") %>%
    as.matrix()

  # Repeat each year's population across 12 months, then / 12
  Ext <- Ext_yr[, rep(seq_len(ncol(Ext_yr)), each = 12)] / 12
  colnames(Ext) <- colnames(Dxt)

  stopifnot(all(dim(Dxt) == dim(Ext)))
  list(Dxt = Dxt, Ext = Ext)
}

# Opioid
hmd_matrices_mo <- list()
for (reg in regions) {
  for (sex in sexes) {
    key <- paste(reg, sex, sep = "_")
    hmd_matrices_mo[[key]] <- make_matrices_monthly(
      wonder_monthly, pop_clean, reg, sex)
  }
}

# Non-opioid
hmd_matrices_mo_no <- list()
for (reg in regions) {
  for (sex in sexes) {
    key <- paste(reg, sex, sep = "_")
    hmd_matrices_mo_no[[key]] <- make_matrices_monthly(
      nonopioid_monthly, pop_clean, reg, sex)
  }
}

cat("Monthly matrices built (opioid + non-opioid)\n")
```

```{r monthly-demogdata}
build_demogdata_monthly <- function(matrices, age_starts, monthly_times, label_suffix = "") {
  result <- list()
  for (reg in regions) {
    mat_m <- matrices[[paste(reg, "Male",   sep = "_")]]
    mat_f <- matrices[[paste(reg, "Female", sep = "_")]]
    lbl <- if (label_suffix == "") reg else paste(reg, label_suffix)
    result[[reg]] <- demogdata(
      data  = mat_m$Dxt / mat_m$Ext,
      pop   = mat_m$Ext,
      ages  = age_starts,
      years = monthly_times,
      type  = "mortality",
      label = lbl,
      name  = "male"
    )
    result[[reg]]$rate$female <- mat_f$Dxt / mat_f$Ext
    result[[reg]]$pop$female  <- mat_f$Ext
  }
  result
}

hmd_region_mo    <- build_demogdata_monthly(hmd_matrices_mo,    age_starts, monthly_times)
hmd_region_mo_no <- build_demogdata_monthly(hmd_matrices_mo_no, age_starts, monthly_times, "non-opioid")
```

```{r monthly-stmomodata}
build_stmomo_monthly <- function(demog_list, cov_df = NULL) {
  stmomo_list <- list()
  for (reg in regions) {
    for (series in c("male", "female")) {
      key <- paste(reg, series, sep = "_")
      stmomo_list[[key]] <- StMoMoData(demog_list[[reg]], series = series, type = "central")
      if (!is.null(cov_df)) {
        cov_reg_mo <- cov_df %>%
          filter(Region == reg) %>% arrange(Year) %>%
          slice(rep(seq_len(n()), each = 12)) %>%
          select(-Region, -Year)
        stmomo_list[[key]]$covariates <- cov_reg_mo
      }
    }
  }
  stmomo_list
}

# Opioid
stmomo_mo     <- build_stmomo_monthly(hmd_region_mo)
stmomo_mo_cov <- build_stmomo_monthly(hmd_region_mo, cov_clean)

# Non-opioid
stmomo_mo_no     <- build_stmomo_monthly(hmd_region_mo_no)
stmomo_mo_no_cov <- build_stmomo_monthly(hmd_region_mo_no, cov_clean)

cat("Monthly StMoMoData created (opioid + non-opioid)\n")
```

# Verification

```{r summary-table}
bind_rows(
  flat_yearly %>% mutate(Type = "Opioid"),
  flat_yearly_nonopioid %>% mutate(Type = "Non-opioid")
) %>%
  group_by(Type, Region, Sex) %>%
  summarise(
    Total_Deaths = sum(Deaths),
    Mean_Rate    = mean(Death_Rate),
    Min_Rate     = min(Death_Rate),
    Max_Rate     = max(Death_Rate),
    .groups = "drop"
  ) %>%
  knitr::kable(digits = 6,
               caption = "Summary by Region and Sex (2005-2019, Ages 15-54)")
```

```{r plot-verify, fig.width=10, fig.height=8}
bind_rows(
  flat_yearly %>% mutate(Type = "Opioid"),
  flat_yearly_nonopioid %>% mutate(Type = "Non-opioid")
) %>%
  ggplot(aes(x = Year, y = log(Death_Rate),
             color = Age_Group, linetype = Sex)) +
  geom_line() +
  facet_grid(Type ~ Region) +
  labs(
    title = "Log Mortality Rate by Region, Age Group, and Sex (2005-2019)",
    y     = "Log(Death Rate)",
    color = "Age Group"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Save Outputs

```{r save}
out <- "../Data/wonder_opioid"

# --- Opioid ---
# Yearly
saveRDS(stmomo_yr, file.path(out, "hmd_yearly_no_cov.rds"))
saveRDS(stmomo_yr_cov, file.path(out, "hmd_yearly_with_cov.rds"))
saveRDS(hmd_region_yr, file.path(out, "demogdata_yearly.rds"))
saveRDS(flat_yearly, file.path(out, "flat_yearly.rds"))
# Monthly
saveRDS(stmomo_mo, file.path(out, "hmd_monthly_no_cov.rds"))
saveRDS(stmomo_mo_cov, file.path(out, "hmd_monthly_with_cov.rds"))
saveRDS(hmd_region_mo, file.path(out, "demogdata_monthly.rds"))
saveRDS(flat_monthly, file.path(out, "flat_monthly.rds"))

# --- Non-opioid ---
# Yearly
saveRDS(stmomo_yr_no, file.path(out, "nonopioid_hmd_yearly_no_cov.rds"))
saveRDS(stmomo_yr_no_cov, file.path(out, "nonopioid_hmd_yearly_with_cov.rds"))
saveRDS(hmd_region_yr_no, file.path(out, "nonopioid_demogdata_yearly.rds"))
saveRDS(flat_yearly_nonopioid, file.path(out, "nonopioid_flat_yearly.rds"))
# Monthly
saveRDS(stmomo_mo_no, file.path(out, "nonopioid_hmd_monthly_no_cov.rds"))
saveRDS(stmomo_mo_no_cov, file.path(out, "nonopioid_hmd_monthly_with_cov.rds"))
saveRDS(hmd_region_mo_no, file.path(out, "nonopioid_demogdata_monthly.rds"))
saveRDS(flat_monthly_nonopioid, file.path(out, "nonopioid_flat_monthly.rds"))

cat("Saved:\n")
cat("  Opioid — Yearly:\n")
cat("    hmd_yearly_no_cov.rds   - StMoMoData [8 ages x 15 years]\n")
cat("    hmd_yearly_with_cov.rds - StMoMoData + covariates\n")
cat("    demogdata_yearly.rds    - demogdata objects\n")
cat("    flat_yearly.rds         - flat data frame\n")
cat("  Opioid — Monthly:\n")
cat("    hmd_monthly_no_cov.rds   - StMoMoData [8 ages x 180 months]\n")
cat("    hmd_monthly_with_cov.rds - StMoMoData + covariates\n")
cat("    demogdata_monthly.rds    - demogdata objects\n")
cat("    flat_monthly.rds         - flat data frame\n")
cat("  Non-opioid — Yearly:\n")
cat("    nonopioid_hmd_yearly_no_cov.rds   - StMoMoData [8 ages x 15 years]\n")
cat("    nonopioid_hmd_yearly_with_cov.rds - StMoMoData + covariates\n")
cat("    nonopioid_demogdata_yearly.rds    - demogdata objects\n")
cat("    nonopioid_flat_yearly.rds         - flat data frame\n")
cat("  Non-opioid — Monthly:\n")
cat("    nonopioid_hmd_monthly_no_cov.rds   - StMoMoData [8 ages x 180 months]\n")
cat("    nonopioid_hmd_monthly_with_cov.rds - StMoMoData + covariates\n")
cat("    nonopioid_demogdata_monthly.rds    - demogdata objects\n")
cat("    nonopioid_flat_monthly.rds         - flat data frame\n")
cat("\nCovariate columns:\n")
print(names(stmomo_yr_cov[["Northeast_male"]]$covariates))
```
