---
title: "Combine Data to HMD Format"
author: "PSTAT 296A"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, message=FALSE, warning=FALSE}
rm(list=ls());gc()
library(tidyverse)
library(demography)
library(StMoMo)
```

# Load Data

```{r load-data}
# Deaths from WONDER (monthly, by region/sex/age group)
wonder <- read.csv("../Data/missing_data/Inferred-WONDER-Census Region-Sex-5Yr.csv")

# Population by region, age, sex (for age-specific exposures)
pop <- read.csv("../Data/covariate/population_by_region_age_sex.csv")

# State-level total population (for regional total population covariate)
state_pop <- read.csv("../Data/covariate/state_population.csv")

# Covariates by region and year
covariates <- read.csv("../Data/covariate/covariate_combine_region_pre_whitening.csv")

covariates <- covariates |> select(-rate_health_care_coverage, -rate_alochol_consum, - rate_health_care_coverage_pw, -rate_alochol_consum_pw, - socio_score_pw, - socio_score_ccf, - opioid_deaths_pw)
```

# Data Cleaning and Harmonization

The three data sources use different naming conventions for Census Regions:

- WONDER: `"Census Region 1: Northeast"`, `"Census Region 2: Midwest"`, etc.
- Population: `"North Central"` (= Midwest), `"Northeast"`, `"South"`, `"West"`
- Covariates: `"Midwest"`, `"Northeast"`, `"South"`, `"West"`

We standardize all to: **Northeast, Midwest, South, West**.

```{r clean-data}
# Define target age groups (5-year bins, ages 15-54)
age_groups <- c("15-19", "20-24", "25-29", "30-34",
                "35-39", "40-44", "45-49", "50-54")
age_starts <- c(15, 20, 25, 30, 35, 40, 45, 50)
age_map    <- setNames(age_starts, age_groups)

target_years <- 2005:2019

# ---- Clean WONDER deaths ----
wonder_clean <- wonder %>%
  mutate(
    Region = case_when(
      Census.Region == "Census Region 1: Northeast" ~ "Northeast",
      Census.Region == "Census Region 2: Midwest"   ~ "Midwest",
      Census.Region == "Census Region 3: South"     ~ "South",
      Census.Region == "Census Region 4: West"      ~ "West"
    ),
    Deaths = suppressWarnings(as.numeric(Deaths))
  ) %>%
  filter(
    Year %in% target_years,
    Age.Group.Code %in% age_groups,
    Sex.Code %in% c("M", "F")
  ) %>%
  # Aggregate monthly to yearly
  group_by(Region, Sex.Code, Age.Group.Code, Year) %>%
  summarise(
    n_months = sum(!is.na(Deaths)),  # track months with non-suppressed data
    Deaths   = sum(Deaths, na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  mutate(
    Age = age_map[Age.Group.Code],
    Sex = ifelse(Sex.Code == "M", "Male", "Female")
  )

# ---- Clean Population (exposures) ----
pop_clean <- pop %>%
  mutate(
    Region = ifelse(census_region == "North Central", "Midwest", census_region)
  ) %>%
  filter(
    year %in% target_years,
    age5 %in% age_groups,
    sex %in% c("Male", "Female")
  ) %>%
  mutate(Age = age_map[age5]) %>%
  rename(Year = year, Sex = sex, Population = population)

# ---- Check for suppressed months ----
suppressed_summary <- wonder_clean %>% filter(n_months < 12)
if (nrow(suppressed_summary) > 0) {
  cat("WARNING: Some year/region/sex/age combinations had months with suppressed data:\n")
  suppressed_summary %>%
    group_by(Region, Sex, Age.Group.Code) %>%
    summarise(
      n_years_affected = n(),
      avg_months_observed = mean(n_months),
      .groups = "drop"
    ) %>%
    print(n = 50)
} else {
  cat("All combinations have 12 months of data. No suppression issues.\n")
}
```

# Create HMD-Format Matrices

For each **Region x Sex** combination, we create:

- `Dxt`: death counts matrix (8 age groups x 15 years)
- `Ext`: exposure/population matrix (8 age groups x 15 years)

This matches the `StMoMoData` format used in `hmd_analysis.Rmd`.

```{r create-matrices}
regions <- c("Northeast", "Midwest", "South", "West")
sexes   <- c("Male", "Female")

# Function to create Dxt and Ext matrices
make_matrices <- function(deaths_df, pop_df, region, sex) {
  Dxt <- deaths_df %>%
    filter(Region == region, Sex == sex) %>%
    select(Age, Year, Deaths) %>%
    pivot_wider(names_from = Year, values_from = Deaths) %>%
    arrange(Age) %>%
    column_to_rownames("Age") %>%
    as.matrix()

  Ext <- pop_df %>%
    filter(Region == region, Sex == sex) %>%
    select(Age, Year, Population) %>%
    pivot_wider(names_from = Year, values_from = Population) %>%
    arrange(Age) %>%
    column_to_rownames("Age") %>%
    as.matrix()

  # Validate
  stopifnot(all(dim(Dxt) == dim(Ext)))
  stopifnot(all(rownames(Dxt) == rownames(Ext)))
  stopifnot(all(colnames(Dxt) == colnames(Ext)))

  list(Dxt = Dxt, Ext = Ext)
}

# Build all matrices
hmd_matrices <- list()
for (reg in regions) {
  for (sex in sexes) {
    key <- paste(reg, sex, sep = "_")
    hmd_matrices[[key]] <- make_matrices(wonder_clean, pop_clean, reg, sex)
    cat(sprintf("  %s: Dxt[%d x %d]\n", key,
                nrow(hmd_matrices[[key]]$Dxt), ncol(hmd_matrices[[key]]$Dxt)))
  }
}
```

# Version 1: HMD Format Without Covariates

We create `demogdata` objects (compatible with `demography` and `StMoMo` packages) for each
region, containing both male and female mortality series. Then convert to `StMoMoData`.

```{r version1-demogdata}
# Create demogdata objects for each region
hmd_region <- list()

for (reg in regions) {
  mat_m <- hmd_matrices[[paste(reg, "Male",   sep = "_")]]
  mat_f <- hmd_matrices[[paste(reg, "Female", sep = "_")]]

  # Central death rates: mx = Dx / Ex
  rate_male   <- mat_m$Dxt / mat_m$Ext
  rate_female <- mat_f$Dxt / mat_f$Ext

  # Create demogdata with male series
  hmd_region[[reg]] <- demogdata(
    data  = rate_male,
    pop   = mat_m$Ext,
    ages  = age_starts,
    years = target_years,
    type  = "mortality",
    label = reg,
    name  = "male"
  )

  # Add female series
  hmd_region[[reg]]$rate$female <- rate_female
  hmd_region[[reg]]$pop$female  <- mat_f$Ext
}

cat("demogdata objects created for:", paste(names(hmd_region), collapse = ", "), "\n")
```

```{r version1-stmomodata}
# Convert to StMoMoData for each region x sex
stmomo_data <- list()

for (reg in regions) {
  for (series in c("male", "female")) {
    key <- paste(reg, series, sep = "_")
    stmomo_data[[key]] <- StMoMoData(hmd_region[[reg]],
                                      series = series,
                                      type   = "central")
    cat(sprintf("  StMoMoData %-20s | ages: %s | years: %d-%d\n", key,
                paste(stmomo_data[[key]]$ages, collapse = ","),
                min(stmomo_data[[key]]$years),
                max(stmomo_data[[key]]$years)))
  }
}
```

### Quick Verification: Mortality Rates from StMoMoData

```{r verify-v1, fig.width=10, fig.height=6}
# Verify: mortality rate = Dxt / Ext
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
for (reg in regions) {
  key_m <- paste(reg, "male", sep = "_")
  mort  <- stmomo_data[[key_m]]
  mx    <- mort$Dxt / mort$Ext
  matplot(mort$years, t(log(mx)), type = "l", lty = 1,
          main = paste(reg, "- Male log(mx)"),
          xlab = "Year", ylab = "log(mx)",
          col = rainbow(length(mort$ages)))
  legend("topright", legend = paste0(age_groups), col = rainbow(8),
         lty = 1, cex = 0.6, ncol = 2)
}
par(mfrow = c(1, 1))
```

# Version 2: HMD Format With Covariates

We merge covariates into two formats:

1. **Flat data frame**: one row per region/sex/age/year with deaths, population, rate, and covariates
2. **StMoMoData + covariates list**: StMoMoData objects with covariates attached

```{r version2-flat}
# Clean covariates
cov_clean <- covariates %>%
  rename(Region = Census_Region) %>%
  filter(Year %in% target_years)

# ---- Flat data frame ----
hmd_flat <- wonder_clean %>%
  select(Region, Sex, Age, Age.Group.Code, Year, Deaths, n_months) %>%
  left_join(
    pop_clean %>% select(Region, Sex, Age, Year, Population),
    by = c("Region", "Sex", "Age", "Year")
  ) %>%
  mutate(Rate = Deaths / Population) %>%
  left_join(cov_clean, by = c("Region", "Year"))

cat(sprintf("Flat data: %d rows x %d columns\n", nrow(hmd_flat), ncol(hmd_flat)))
cat("Columns:", paste(names(hmd_flat), collapse = ", "), "\n")
```

```{r version2-stmomo-cov}
# ---- StMoMoData + covariates list ----
stmomo_data_cov <- list()

for (reg in regions) {
  cov_reg <- cov_clean %>%
    filter(Region == reg) %>%
    arrange(Year) %>%
    select(-Region)

  for (series in c("male", "female")) {
    key <- paste(reg, series, sep = "_")

    stmomo_data_cov[[key]] <- stmomo_data[[key]]
    stmomo_data_cov[[key]]$covariates <- cov_reg
  }
}

cat("StMoMoData + covariates created for:", paste(names(stmomo_data_cov), collapse = ", "), "\n")
cat("\nCovariate columns per region:\n")
print(names(stmomo_data_cov[["Northeast_male"]]$covariates))
```

# Add Regional Total Population

Aggregate state-level population to Census Regions and add as an additional covariate.

```{r regional-total-pop}
state_region_map <- tibble(
  state_name = c(
    # Northeast
    "Connecticut", "Maine", "Massachusetts", "New Hampshire", "New Jersey",
    "New York", "Pennsylvania", "Rhode Island", "Vermont",
    # Midwest
    "Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota",
    "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin",
    # South
    "Alabama", "Arkansas", "Delaware", "District of Columbia", "Florida",
    "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi",
    "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas",
    "Virginia", "West Virginia",
    # West
    "Alaska", "Arizona", "California", "Colorado", "Hawaii", "Idaho",
    "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming"
  ),
  Region = c(
    rep("Northeast", 9), rep("Midwest", 12),
    rep("South", 17),    rep("West", 13)
  )
)

regional_pop <- state_pop %>%
  filter(year %in% target_years) %>%
  left_join(state_region_map, by = "state_name") %>%
  group_by(Region, year) %>%
  summarise(Total_Population = sum(population), .groups = "drop") %>%
  rename(Year = year)

# Add to flat data
hmd_flat <- hmd_flat %>%
  left_join(regional_pop, by = c("Region", "Year"))

cat("Final flat data columns:\n")
str(hmd_flat, give.attr = FALSE)
```

# Verification

```{r summary-table}
hmd_flat %>%
  group_by(Region, Sex) %>%
  summarise(
    Total_Deaths  = sum(Deaths),
    Mean_Rate     = mean(Rate),
    Min_Rate      = min(Rate),
    Max_Rate      = max(Rate),
    .groups = "drop"
  ) %>%
  knitr::kable(digits = 6, caption = "Summary by Region and Sex (2005-2019, Ages 15-54)")
```

```{r plot-verify, fig.width=10, fig.height=8}
hmd_flat %>%
  ggplot(aes(x = Year, y = log(Rate), color = Age.Group.Code, linetype = Sex)) +
  geom_line() +
  facet_wrap(~Region, ncol = 2) +
  labs(
    title = "Log Mortality Rate by Region, Age Group, and Sex (2005-2019)",
    y     = "Log(Central Death Rate)",
    color = "Age Group"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Save Outputs

```{r save}
# Version 1: Without covariates
saveRDS(stmomo_data, file = "../Data/wonder_opioid/hmd_format_no_cov.rds")
saveRDS(hmd_region,  file = "../Data/wonder_opioid/demogdata_by_region.rds")

# Version 2: With covariates
saveRDS(stmomo_data_cov, file = "../Data/wonder_opioid/hmd_format_with_cov.rds")
write.csv(hmd_flat, file = "../Data/wonder_opioid/hmd_flat_with_covariates.csv", row.names = FALSE)

cat("Saved outputs:\n")
cat("  Version 1 (no covariates):\n")
cat("    ../Data/hmd_format_no_cov.rds    - list of 8 StMoMoData objects (4 regions x 2 sexes)\n")
cat("    ../Data/demogdata_by_region.rds  - list of 4 demogdata objects (male + female per region)\n")
cat("  Version 2 (with covariates):\n")
cat("    ../Data/hmd_format_with_cov.rds  - list of 8 StMoMoData objects + $covariates\n")
cat("    ../Data/hmd_flat_with_covariates.csv - flat data frame for analysis\n")
```
